---
title: "processNDVI"
output: html_document
---

This code opens up the NDVI files created / collected by Shelley Petroy on Geotiff
format, and crops them using a shapefile created by leah wasser. Then creates an output geotiff.

bla bla bla

```{r}

library(raster)
library(ggplot2)
library(rgdal)

#set WD
setwd("~/Documents/data/1_DataPortal_Workshop")

#define working directory
#Contains Harvard D01
workingDir <- "/Volumes/data-spetroy/Landsat Domain 1/LS5/P12R31/2011/"
#D17 data
#workingDir <- "/Volumes/data-spetroy/Landsat Domain 17/P43R34/2011/"

#Landsat Domain 17/P43R34/2011/

#define Year
#could pull this from the path possibly?
yr <- 2011
#site <- "SJER"
site <- "HARV"

#Name of clipping file
#note - clip using a file in the same PROJ as the landsat time series. 
#this is not always the BEST CRS for that particular location given 
#landsat scenes cross UTM zones.
clipFilePath <- "NDVI/"
#clipFileName <- "SJERClip_UTM10N"
#use zone 19N for landsat
clipFileName <- "HarvardClipBox"


#create list of files to make raster stack
rasterlist <-  list.files(workingDir, full.names=TRUE, pattern = "ndvi.tif$")

#read in the shapefile
#squarePlot <- readShapePoly(clipFile)
squarePlot <- readOGR(clipFilePath,clipFileName)

#define the path to write tiffs
#tifPath <- "Landsat_TimeSeries/D01/LS5/P12R31/2011/"
#SJER path
#tifPath <- "NDVI/SJER/2011/"
#Harvard path
tifPath <- "NDVI/HARV/2011/"

#generate the directory to write out tiffs


for (tifFile in rasterlist) {
  #read in tiff as raster
  image_R <- raster(tifFile)
  #grab the file name
  a <- names(image_R)
  
  #then pull out the julian date as a name
  a <- substr(a,nchar(a)-15,nchar(a)-13)
  
  #clean up periods
  #a <- gsub(".","_",a,fixed=TRUE)
  #add file extention
  b <- paste(a,site,"ndvi_crop.tiff", sep="_")
  
  #create the final path
  c <- paste(tifPath,b,sep="")
  
  #crop the raster
  imageCrop <- crop(image_R, squarePlot)

  #write to geotiff
  writeRaster(imageCrop,c,"GTiff", overwrite = TRUE)
}


```

Next let's plot the data.

```{r}
#open up the cropped files
#create list of files to make raster stack
allCropped <-  list.files(tifPath, full.names=TRUE)

#create a raster stack from the list
rastStack <- stack(allCropped)
plot(rastStack, zlim=c(1500,10000))
hist(rastStack,xlim=c(1500,10000))
#create dataframe for NDVI data by Julian Day
m <- as.data.frame(matrix(-999, ncol = 2, nrow = length(allCropped)))
colnames(m) <- c("julianDays", "meanNDVI")
i <- 0
for (crop in allCropped){
  i=i+1
  #open raster
  imageCrop <- raster(crop)
  
  #calculate the mean of each
  m$meanNDVI[i] <- cellStats(imageCrop,mean) 
  
  #grab julian days
  #m$julianDays[i] <- substr(crop,nchar(crop)-16,nchar(crop)-14)
  #new naming convention
  m$julianDays[i] <- substr(crop,nchar(crop)-21,nchar(crop)-19)
  
  m$year <- yr
  m$site <- site
}

##plot stuff
qplot(m$julianDays,m$meanNDVI)

ggplot(m, aes(julianDays, meanNDVI)) +
  geom_point(size=4,colour = "blue") + 
  ggtitle("NDVI for 2011 SJER\nLandsat Derived") +
  xlab("Julian Days") + ylab("Mean NDVI") +
  theme(text = element_text(size=20)) +
  ylim(c(0,8000))

```


```{r}
#create animation ot the NDVI outputs
library(animation)

#if(!file.exists("ndvi.gif")) { # Check if the file exists
  saveGIF(
    for (i in 1:length(allCropped)) {
                      plot(rastStack[[i]],
                      main=names(rastStack[[i]]),
                      legend.lab="NDVI",
                      col=rev(terrain.colors(30)),
                      zlim=c(1500,10000) )
      }, 
    movie.name = "ndvi.gif", 
    ani.width = 300, ani.height = 300, 
    interval=.5)
#}

    
```

#CREATE RGB files

The code below will take an input EXTENT shapefile in the SAME PROJ as the landsat
imagery and will clip the R,G and B band stack to create an output RGB image for each 
geotiff available in the directory. It looks for unique scenes by Julian Day.

Note that you must input the site acronym and clipping polygon manually at the 
top of the code chunk. Also note that the extent for the RGB imagery is LARGER
than the NDVI chunks. This is because the data are 30 m and the premise is that 
we are taking an area that is close to the tower. But a few landsat pixels don't 
render in RGB in a visual way that looks like an image (it looks pixelated) so we 
chose a broader area for the RGB imagery.

We could clip NDVI to the larger area but the analysis will begin to be less useful
given it's a average over a larger area.

## About the data
The data are pulled from hte drive where Sheley collected the 30 year time series
for landsat. We could also pull a fourth NIR band to allow students to create
NDVI which might be interesting.

#Clip the landsat data to a ROI that allows students to see the area and the clouds

```{r clip-landsat-create-rgb }
########### define Year
#could pull this from the path possibly?
yr <- 2011
  
#DEFINE SITE
site <- "SJER"
#site <- "HARV"

#Contains SJER D17
workingDir <- "/Volumes/data-spetroy/Landsat Domain 17/P43R34/2011/"

#Contains Harvard D01
#workingDir <- "/Volumes/data-spetroy/Landsat Domain 1/LS5/P12R31/2011/"

##### DEFINE BOUNDARY FILES FOR CLIPPING & DIR
#for HARV
#clipFileName <- "landClip_UTMZ19N"
#clipFilePath <- "boundaryFiles/HARV/"
#for SJER
#clipFileName <- "SJERClip_UTM10N"
#the file for the RGB extent so the image looks more like an image and less pixelated
clipFileName <- "SJERClipRGB_UTM10N"
clipFilePath <- "boundaryFiles/SJER/"



createLandRGB <- function(aPath,jDday){

#this function will look through a dir and create an RGB image

}
  
#create list of files to make raster stack
#find only files that end with "tif" and are band files
bandList <-  list.files(workingDir, full.names=TRUE, pattern = "01_B|00_B.*TIF$")
  
#get path start name
#tPath <- substr(bandList[1],0,nchar(i)-15)

#make a list of UNIQUE julian day values
j <- 0
jDay <- 0
for (i in bandList) {
  j <- j+1
  #get julian day from path  
  jDay[[j]] <- substr(i,nchar(i)-14,nchar(i)-12)
  jDayUn <- unique(jDay)
}



#clip stack
#squarePlot <- readShapePoly(clipFile)
landClip <- readOGR(clipFilePath,clipFileName)
#define bands
band1 <- 1
band2 <- 2
band3 <- 3

aDir <- "RGB"

#loop through julian days and process landsat image
for (k in jDayUn){
  
  #filter files that contain paste0("1",k)
  #file the files that contain the first Julian Day Value, K
  bFiles <- bandList[grepl(paste0("1",k),bandList)]
  
  #create basePath
  basePath <- substr(bFiles[1],0,nchar(bFiles[1])-5)
  
  #create paths
  band1Path <- paste0(basePath,band1,".TIF", collapse = "")
  #create paths
  band2Path <- paste0(basePath,band2,".TIF", collapse = "")
  #create paths
  band3Path <- paste0(basePath,band3,".TIF", collapse = "")
  #processLandsat
  #create raster stack
  newStack <- stack(band3Path,band2Path,band1Path)
  
  #clip raster stack
  cropStack <- crop(newStack,landClip)
  
  #create tif name
  #landName <- paste0(k,"_HARV_landRGB.tif")
  landName <- paste0(k,"_SJER_landRGB.tif")
  
  #tifPath
  tifPath <- paste0("Landsat_NDVI/",site,"/",yr,"/",aDir,"/",landName)
  
  #writeGeotiff
  writeRaster(cropStack,tifPath,"GTiff", overwrite = TRUE)
  
}

#read in file
new <- stack("Landsat_NDVI/HARV/2011/RGB/309_HARV_landRGB.tif")
plot(new[1])
plotRGB(new, r=1,g=2,b=3, stretch="lin")  

#read in file
new <- stack("Landsat_NDVI/HARV/2011/RGB/293_HARV_landRGB.tif")

plotRGB(new, r=1,g=2,b=3, stretch="lin")  
  
```

#The code below was use
to crop the AOP data used in the raster lessons
```{r clip-aop-data }

library(raster)
library(rgdal)
#set WD
setwd("~/Documents/data/1_DataPortal_Workshop")

clipRaster <- function(rPath,shpPath,shpName,writeName){
  #a small function that inputs a raster and a crop shapefile 
  #and outputs a cropped raster
  
  #read in raster
  rast <- raster(rPath)

  #read clip file
  AOPClip <- readOGR(shpPath,shpName)

  #crop the raster
  rCrop <- crop(rast,AOPClip)

  #write geotiff
  writeRaster(rCrop,writeName,"GTiff", overwrite = TRUE, NAflag=-9999)
}

#clipDSM
clipRaster("NEON_RemoteSensing/HARV/DSM/2014_HARV_2_DSM.tif",
           "boundaryFiles/HARV/",
           "AOPClip_UTMz18N",
           "NEON_RemoteSensing/HARV/DSM/HARV_dsmCrop.tif")
#clip CHM
clipRaster("NEON_RemoteSensing/HARV/CHM/2014_HARV_2_CHM.tif",
           "boundaryFiles/HARV/",
           "AOPClip_UTMz18N",
           "NEON_RemoteSensing/HARV/CHM/HARV_chmCrop.tif")

#clip DTM
clipRaster("NEON_RemoteSensing/HARV/DTM/2014_HARV_2_DTM.tif",
           "boundaryFiles/HARV/",
           "AOPClip_UTMz18N",
           "NEON_RemoteSensing/HARV/DTM/HARV_dtmCrop.tif")


########## CROP SJER Data
#clip DTM
clipRaster("NEON_RemoteSensing/SJER/DTM/SJER2013_DTM.tif",
           "boundaryFiles/SJER/",
           "SJER_AOPClip_UTMZ11n",
           "NEON_RemoteSensing/SJER/DTM/SJER_dtmCrop.tif")

#clip DSM
clipRaster("NEON_RemoteSensing/SJER/DSM/SJER2013_DSM.tif",
           "boundaryFiles/SJER/",
           "SJER_AOPClip_UTMZ11n",
           "NEON_RemoteSensing/SJER/DSM/SJER_dsmCrop.tif")

#http://www.r-forge.r-project.org/forum/forum.php?thread_id=3061&forum_id=962&group_id=294

chm <- raster("NEON_RemoteSensing/HARV/CHM/2014_HARV_2_CHM.tif")

chm@file@nodatavalue

NAvalue(chm)
```