---
layout: post
title: "Lesson 01: Plot Raster Data in R"
date:   2015-10-28
authors: [Kristina Riemer, Jason Williams, Jeff Hollister, Mike Smorul, 
Zack Brym, Leah Wasser]
contributors: [Megan A. Jones]
packagesLibraries: [raster, rgdal]
dateCreated:  2015-10-23
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
category: spatio-temporal-workshop
tags: [raster-ts-wrksp, raster]
mainTag: raster-ts-wrksp
description: "This lesson reviews how to plot a raster in R using the plot() 
command. It also covers how to layer a raster on top of a hillshade for a 
eloquent map.
output."
code1: Plot-Rasters-In-R.R
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: http://www.neoninc.org
permalink: /R/Plot-Rasters-In-R
comments: false
---

{% include _toc.html %}

##About
This lesson reviews how to use the `plot` function in `R` to plot raster data.

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

###Goals / Objectives

After completing this activity, you will:

* Be able how to plot a single band raster in R.
* Be able to layer a raster dataset on top of a hillshade to create an elegant 
basemap.

###Challenge Code
Throughout the lesson we have Challenges that reinforce learned skills. Possible
solutions to the challenges are not posted on this page, however, the code for 
each challenge is in the `R` code that can be downloaded for this lesson (see 
footer on this page).

###Things You'll Need To Complete This Lesson
You will need the most current version of R, and preferably RStudio, loaded on
your computer to complete this lesson.

###R Libraries to Install:

* **raster:** `install.packages("raster")`
* **rgdal:** `install.packages("rgdal")`

####Data to Download

Download the raster files for the Harvard Forest dataset:

<a href="http://files.figshare.com/2434040/NEON_RemoteSensing.zip" class="btn btn-success"> DOWNLOAD Sample NEON Airborne Observation Platform Raster Data</a> 

The LiDAR and imagery data used to create the rasters in this dataset were 
collected over the <a href="http://www.neoninc.org/science-design/field-sites/harvard-forest" target="_blank" >Harvard</a> and 
<a href="http://www.neoninc.org/science-design/field-sites/san-joaquin-experimental-range" target="_blank" >San Joaquin</a> field sites 
and processed at <a href="http://www.neoninc.org" target="_blank" >NEON </a> 
headquarters. The entire dataset can be accessed by request from the 
<a href="http://www.neoninc.org/data-resources/get-data/airborne-data" target="_blank"> NEON 
website.</a>

####Setting the Working Directory
The code in this lesson assumes that you have set your working directory to the
location of the unzipped file of data downloaded above.  If you would like a
refresher on setting the working directory, please view the [Setting A Working Directory In R]({{site.baseurl}}/R/Set-Working-Directory "R Working Directory Lesson") 
lesson prior to beginning this lesson.

####Raster Lesson Series 
This lesson is a part of a series of raster data in R lessons:

* [Lesson 00 - Intro to Raster Data in R]({{ site.baseurl}}/R/Introduction-to-Raster-Data-In-R/)
* [Lesson 01 - Plot Raster Data in R]({{ site.baseurl}}/R/Plot-Rasters-In-R/)
* [Lesson 02 - Reproject Raster Data in R]({{ site.baseurl}}/R/Reproject-Raster-In-R/)
* [Lesson 03 - Raster Calculations in R]({{ site.baseurl}}/R/Raster-Calculations-In-R/)
* [Lesson 04 - Work With Multi-Band Rasters - Images in R]({{ site.baseurl}}/R/Multi-Band-Rasters-In-R/)
* [Lesson 05 - Raster Time Series Data in R]({{ site.baseurl}}/R/Raster-Times-Series-Data-In-R/)
* [Lesson 06 - Plot Raster Time Series Data in R Using RasterVis and LevelPlot]({{ site.baseurl}}/R/Plot-Raster-Times-Series-Data-In-R/)
* [Lesson 07- Extract NDVI Summary Values from a Raster Time Series]({{ site.baseurl}}/R/Extract-NDVI-From-Rasters-In-R/)

###Sources of Additional Information

* <a href="http://cran.r-project.org/web/packages/raster/raster.pdf" target="_blank"> Read more about the `raster` package in R.</a>

</div>

##Plot Raster Data in R
In this lesson, we will plot the raster for the Digital Surface Model for
Harvard Forest. We will use the `hist()` function as a tool to explore raster 
values. And render categorical plots, using `breaks` or bins that are meaningful
in our data. 

We will continue to use the `raster` and `rgdal` libraries in this
lesson.  If you do not have the `DSM_HARV` object from [the Intro To Raster In R lesson]({{ site.baseurl}}/R/Introduction-to-Raster-Data-In-R "First Lesson in Series"), please create it now.  

```{r load-libraries }
#if they are not already loaded
library(rgdal)
library(raster)

#set working directory to ensure R can find the file we wish to import
#setwd("working-dir-path-here")

#import raster
DSM_HARV <- raster("NEON_RemoteSensing/HARV/DSM/HARV_dsmCrop.tif")

```

First, let's plot our Digital Surface Model object (`DSM_HARV`) using the `plot`
function. We add a title using `main=""`.

```{r hist-raster }
#Plot raster object
plot(DSM_HARV,
     main="NEON Digital Surface Model\nHarvard Forest Field Site")

```

# Plotting Data Using Breaks
Often we may want to view our data subsetted by a range of values rather than
using a continuous color ramp. This is comparable to a "classified" map in a
tool like ArcGIS or QGIS.

Unless here is a specific reason for having specific breaks, we want to use a
histogram of our data to determine where the breaks in our data should go. We 
can use `breaks=` to tell `R` to use fewer or more breaks or bins in our 
histogram. 

```{r create-histogram-2-breaks }
#Plot distribution of raster values 
hist(DSM_HARV,
     breaks=2,
     main="Histogram Digital Surface Model\nHarvard Forest Field Site",
     col="wheat3")

```

Remember, the default for the histogram is to include only 100,000 values from
the data set.  We could force it to show all the pixel values or we can use the
histogram as is and figure that the sample of 100,000 values represents our data 
well. 

Looking at our histogram, R has binned out the data as follows:

* 300-350m, 350-400m, 400-450m

We have most of our pixel values in the middle range with a few lower values,
and very few higher values.  

We can use those bins to plot our raster data. We will use the 
`terrain.colors(3)` function to create a palette of 3 colors to use in our plot.

We can use the `breaks` method to add breaks.  To specify where the breaks occur,
we need to use the following syntax: `breaks=c(value1,value2,value3)`.

We can include as few or many breaks as we'd like.

```{r plot-with-breaks }
#plot using breaks.
plot(DSM_HARV, 
     breaks = c(350, 300, 400, 450), 
     col = terrain.colors(3),
     main="NEON Digital Surface Model\nHarvard Forest Field Site")

```

If we know how many breaks we want but don't care the exact values where the 
breaks occur we can simply use `breaks=3`. 
```{r plot-with-breaks2 }
#plot using breaks.
plot(DSM_HARV, 
     breaks = 3, 
     col = terrain.colors(3),
     main="NEON Digital Surface Model\nHarvard Forest Field Site")

```

What is the difference between these two plots?  

Why do we only need `breaks =3` but had to specify four values?  

###Formatting Plot Axes
If we're going to be creating multiple plots and we want to easily be able to
change the color across the plots, we can create an `R` object (`myCol`) for the
color we want and use that object in the plot code. Then if we want to change
the color theme (or number of bins) we can just change the code once, not in all
plots.  We can label the x- and y-axes of our plot too. 

```{r add-plot-title }
#Assign color to a object for repeat use/ ease of changing
myCol = terrain.colors(3)

#Add axis labels
plot(DSM_HARV, 
     breaks = c(300, 350, 400, 450), 
     col = myCol,
     main="NEON Digital Surface Model\nHarvard Forest Field Site", 
     xlab = "X Coordinates (m)", 
     ylab = "Y Coordinates (m)")
```

Or we can also turn off the axes altogether. 

```{r turn-off-axes }
#or we can turn off the axis altogether
plot(DSM_HARV, 
     breaks = c(300, 350, 400, 450), 
     col = myCol,
     main="NEON Digital Surface Model\nHarvard Forest Field Site", 
     axes=FALSE)

```

>##Challenge
>
>Create a plot of the Harvard DSM that has a legend with six colors that are 
>evenly divided amongst the range of pixel values. 

``` {r challenge-code-plotting, echo=FALSE }



```

#Layering One Raster on Top of Another 

Sometimes we want to layer a raster on top of another raster and use a transparency
to give it texture. The most common approach to this is to use a `HILLSHADE` as
the base raster layer. A hillshade is simply a raster that maps the shadows and 
texture that you would see from above when viewing terrain. It gives the data a 
nice 3D appearance. 

```{r hillshade }
#import DSM hillshade
hill_HARV <- raster("NEON_RemoteSensing/HARV/DSM/HARV_DSMhill.tif")

#plot hillshade using a grayscale color ramp that looks like shadows.
plot(hill_HARV,
    col=grey(1:100/100),  #create a color ramp of grey colors
    legend=F,
    main="NEON Hillshade - DSM\n Harvard Forest",
    axes=FALSE)

```

Now that we have the hillshade we might then overlay another raster on top by
using `add=TRUE`. Let's overlay `DSM` on top of a `hillshade` of the DSM.

``` {r overlay-hillshade}

#plot hillshade using a grayscale color ramp that looks like shadows.
plot(hill_HARV,
    col=grey(1:100/100),  #create a color ramp of grey colors
    legend=F,
    main="NEON Hillshade - DSM\n Harvard Forest",
    axes=FALSE)

#add the DSM on top of the hillshade
plot(DSM_HARV,
     col=rainbow(100),
     alpha=0.4,
     add=T,
     legend=F)
```

Note that here we used the color palete `rainbow()` instead of `terrain.color()`.

>##Challenge
>Use the files in the `NEON_RemoteSensing/SJER/` to create a map of the SJER DTM
>layered on top of the SJER DTM hillshade. Then create a map of the SJER DSM layered
>on top of the SJER DSM. Experiment with various <a href="https://stat.ethz.ch/R-manual/R-devel/library/grDevices/html/palettes.html" target="_blank">R color palettes</a>
>when you make your maps.

```{r challenge-plots, echo=FALSE}
#import DSM 
DSM_SJER <- raster("NEON_RemoteSensing/SJER/DSM/SJER_DSMcrop.tif")
#import DSM hillshade
hill_SJER <- raster("NEON_RemoteSensing/SJER/DSM/SJER_DSMhill.tif")

#plot hillshade using a grayscale color ramp that looks like shadows.
plot(hill_SJER,
    col=grey(1:100/100),  #create a color ramp of grey colors
    legend=F,
    main="NEON Hillshade - DSM\n SJER Field Site",
    axes=FALSE)

#add the DSM on top of the hillshade
plot(DSM_SJER,
     col=terrain.colors(100),
     alpha=0.4,
     add=T,
     legend=F)

#import DSM 
DTM_SJER <- raster("NEON_RemoteSensing/SJER/DTM/SJER_DTMcrop.tif")
#import DSM hillshade
hillDTM_SJER <- raster("NEON_RemoteSensing/SJER/DTM/SJER_DTMhill.tif")

#plot hillshade using a grayscale color ramp that looks like shadows.
plot(hillDTM_SJER,
    col=grey(1:100/100),  #create a color ramp of grey colors
    legend=F,
    main="NEON Hillshade - Digital Terrain Model\n SJER Field Site",
    axes=FALSE)

#add the DSM on top of the hillshade
plot(DTM_SJER,
     col=terrain.colors(100),
     alpha=0.4,
     add=T,
     legend=F)

```

