---
layout: post
title: "Lesson 01: Plot Rasters in R"
date:   2015-10-28
authors: [Kristina Riemer, Jason Williams, Jeff Hollister, Mike Smorul, 
Zack Brym, Leah Wasser]
contributors: [Megan A. Jones]
packagesLibraries: [raster, rgdal]
dateCreated:  2015-10-23
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
category: spatio-temporal-workshop
tags: [module-1]
mainTag: GIS-Spatial-Data
description: "This lesson reviews how to plot a raster in R using the plot() command.
It also covers how to overlay a raster on top of a hillshade for a eloquent map
output."
code1: 
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: http://www.neoninc.org
permalink: /R/Plot-Rasters-In-R.R
comments: false
---

{% include _toc.html %}

##About
This post explains the simple plotting function in the `raster` package.

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

###Goals / Objectives

After completing this activity, you will:

* Be able how to plot a single band raster in R.
* Be able to layer rasters to create a plot with hillshading.

###Things You'll Need To Complete This Lesson
You will need the most current version of R, and preferably RStudio, loaded on
your computer to complete this lesson.

###R Libraries to Install:

* **raster:** `install.packages("raster")`
* **rgdal:** `install.packages("rgdal")`

####Data to Download

Download the workshop data:

* <a href="http://figshare.com/articles/NEON_AOP_Hyperspectral_Teaching_Dataset_SJER_and_Harvard_forest/1580086" class="btn btn-success"> DOWNLOAD Sample NEON LiDAR data in Raster Format & Vegetation Sampling Data</a>

The LiDAR and imagery data used to create the rasters in this dataset were 
collected over the Harvard and San Joaquin field sites 
and processed at <a href="http://www.neoninc.org" target="_blank" >NEON </a> 
headquarters. The entire dataset can be accessed by request from the NEON website.

####Recommended Pre-Lesson Reading

* <a href="http://cran.r-project.org/web/packages/raster/raster.pdf" target="_blank">
Read more about the `raster` package in R.</a>

</div>


```{r pseudo-code }
# Plot raster file and change some parameters
plot(DSM) #Necessary to explicitly differentiate between base plot and raster plot?
pixels <- ncol(DSM) * nrow(DSM)
hist(DSM, col = "blue", maxpixels = pixels)
myCol <- terrain.colors(10)
plot(DSM, 
     breaks = c(320, 340, 360, 380, 400), 
     col = myCol,
     maxpixels = pixels) #optional argument

plot(DSM, 
     zlim = c(340, 400)) #optional argument
#`image` v. `plot`
# TODO: challenge

```


##Plotting raster files
We should already be familiar with the DSM Harvard raster file because it has 
been opened in the first raster lesson, Raster Structure. We will now plot this 
raster and the histogram of the raster values to get an idea of what the raster 
is like.  

```{r hist-raster }
#Plot entire raster
plot(DSM)

#Plot distribution of raster values 
hist(DSM)
# Warning? Hist by default only plots subset
pixels = ncol(DSM) * nrow(DSM)
hist(DSM, maxpixels = pixels)
```

We can also use some arguments for the plot function to plot a subset of the 
raster values or choose which values correspond to which plot colors. 

```{r breaks-zlim }
#Plot a subset of the pixel values, e.g., the bottom 50m of surface
plot(DSM, zlim = c(305, 355))

#Plot all pixel values using broader categories for values
myCol = terrain.colors(4)

plot(DSM, 
     breaks = c(305, 341, 377, 416), 
     col = myCol, 
     maxpixels = pixels)
```

Lastly, like most plots, we can add labels.

```{r add-plot-title }
#Add axes and title to raster plot
plot(DSM, 
     xlab = "X Coordinates", 
     ylab = "Y Coordinates", 
     main = "Harvard Forest Digital Surface Model")

```

##Challenge

Create a plot of the Harvard DSM that has a legend with six colors that are 
evenly divided amongst the range of pixel values. 




#Layering rasters
Sometimes you want to layer a raster on to another raster and use a transparency 
to give it texture. The most common approach to this is to use a `HILLSHADE` as
the base raster layer. A hillshade is simply a raster that maps the shadows and 
texture that you would see from above when viewing terrain. It gives data a nice 
3D appearance. 

```{r hillshade }
#import DSM hillshade
hill <- raster("NEON_RemoteSensing/HARV/DSM/HARV_DSMhill.tif")

#	Plot hillshade using a grayscale color ramp that looks like shadows.
plot(hill,
    col=grey(1:100/100),  #
    legend=F,
    main="NEON Hillshade - DSM\n Harvard Forest")
```

Now that we have the hillshade we might then overlay another raster on top by
using `add=TRUE`.
Let's overlay `DSM` on top of a `hillshade` of the DSM.
``` {r overlay-hillshade}

#overlay the DSM on top of the hillshade
plot(DSM,
     col=rainbow(100),
     alpha=0.4,
     add=T,
     legend=F)
```

Here we used the color palete `rainbow()` instead of `terrain.color()`.
