---
layout: post
title: "Raster 02: When Rasters Don't Line Up - Reproject Raster Data in R"
date:   2015-10-27
authors: [Jason Williams, Jeff Hollister, Kristina Riemer, Mike Smorul, Zack Brym, Leah A. Wasser, Megan A. Jones]
contributors: [ ]
packagesLibraries: [raster, rgdal]
dateCreated:  2015-10-23
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
categories:  [self-paced-tutorial]
tags: [R, raster, spatial-data-gis]
workshopSeries: [raster-data]
mainTag: raster-data
description: "This tutorial explores issues associated with working with rasters
in different Coordinate Reference Systems (CRS) / projections. When two rasters 
are in different CRS, they will not plot nicely together on a map. We will learn
how to reproject a raster in R using the projectRaster function in the raster 
package."
code1: 02-Reproject-Raster-In-R.R
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink:
permalink: R/Reproject-Raster-In-R
comments: false
---

{% include _toc.html %}

## About
Sometimes we encounter raster datasets that do not "line up" when plotted or 
analyzed. Rasters that don't line up are most often in different Coordinate
Reference Systems (`CRS`).

This tutorial explains how to deal with rasters in different, known `CRS`s. It
will walk though reprojecting rasters in `R` using the `projectRaster` function
in the `raster` library.

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

# Goals / Objectives

After completing this activity, you will:

* Be able to reproject a raster in R.

## Things You’ll Need To Complete This Lesson
To complete this lesson: you will need the most current version of R, and 
preferably RStudio, loaded on your computer.

### Install R Packages

* **raster:** `install.packages("raster")`
* **rgdal:** `install.packages("rgdal")`

* [More on Packages in R - Adapted from Software Carpentry.]({{site.baseurl}}R/Packages-In-R/)

#### Data to Download
{% include/dataSubsets/_data_Airborne-Remote-Sensing.html %}

****

{% include/_greyBox-wd-rscript.html %}
{% include/tutorialSeries/_series_dc-spatial-raster.html %}

****

### Additional Resources

* <a href="http://cran.r-project.org/web/packages/raster/raster.pdf" target="_blank">
Read more about the `raster` package in `R`.</a>

</div>

## Raster Projection in R

In the [Plot Raster Data in R]({{ site.baseurl}}/R/Plot-Rasters-In-R/) 
lesson, we learned how to layer a raster file on top of a hillshade for a nice
looking basemap. In this lesson, all of our data were in the same `CRS`. What happens when things don't line up?

We will use the `raster` and `rgdal` packages in this lesson.  

```{r load-libraries }

#load raster package
library(raster)
library(rgdal)
```

Let's create an map of the Harvard Forest Digital Terrain Model 
(`DTM_HARV`) draped or layered on top of the hillshade (`DTM_hill_HARV`).

```{r import-DTM-hillshade }
#import DTM
DTM_HARV <- raster("NEON-DS-Airborne-Remote-Sensing/HARV/DTM/HARV_dtmcrop.tif")
#import DTM hillshade
DTM_hill_HARV <- 
  raster("NEON-DS-Airborne-Remote-Sensing/HARV/DTM/HARV_DTMhill_WGS84.tif")

#plot hillshade using a grayscale color ramp 
plot(DTM_hill_HARV,
    col=grey(1:100/100),
    legend=FALSE,
    main="DTM Hillshade\n NEON Harvard Forest Field Site")

#overlay the DTM on top of the hillshade
plot(DTM_HARV,
     col=terrain.colors(10),
     alpha=0.4,
     add=TRUE,
     legend=FALSE)

```

Our results are curious - the Digital Terrain Model (`DTM_HARV`) did not plot on
top of our hillshade. The hillshade plotted just fine on it's own. Let's try to 
plot the DTM on it's own to make sure there are data there.

<i class="fa fa-star"></i> **Data Tip:** For boolean ``R elements, such as
 `add=TRUE`, you can use `T` and `F` in place of `TRUE` and `FALSE`.
{: .notice}

```{r plot-DTM }
#Plot DTM 
plot(DTM_HARV,
     col=terrain.colors(10),
     alpha=1,
     legend=F,
     main="Digital Terrain Model\n NEON Harvard Forest Field Site")

```

Our DTM seems to contain data and plots just fine. Let's next check the
 Coordinate Reference System (`CRS`) and compare it to our hillshade.

```{r explore-crs }
#view crs for DTM
crs(DTM_HARV)

#view crs for hillshade
crs(DTM_hill_HARV)
```

Aha! `DTM_HARV` is in the UTM projection. `DTM_hill_HARV` is in
`Geographic WGS84` - which is represented by latitude and longitude values.
Because the two rasters are in different CRSs, they don't line up when plotted
in `R`. We need to *reproject* `DTM_HARV` into the `UTM CRS`. Alternatively,
we could reproject the hillshade into WGS84. 

## Reproject Rasters
We can use the `projectRaster` function to reproject a raster into a new `CRS`.
Keep in mind that reprojection only works when you first have a *defined* `CRS`
for the raster object that you want to reproject. It cannot be used if *no*
`CRS` is defined. Lucky for us, the `DTM_hill_HARV` has a defined `CRS`. 

To use the `projectRaster` function, we need to define two things:

1. the object we want to reproject and 
2. the CRS that we want to reproject it to. 

The syntax is `projectRaster(RasterObject,crs=CRSToReprojectTo)`

We want the `CRS` of our hillshade to match the `DTM_HARV` raster. We can thus
assign the `CRS` of our `DTM_HARV` to our hillshade within the `projectRaster`
functionas follows: `crs=crs(DTM_HARV)`.

```{r reproject-raster }

#reproject to UTM
DTM_hill_UTMZ18N_HARV <- projectRaster(DTM_hill_HARV, 
                                       crs=crs(DTM_HARV))

#compare attributes of DTM_hill_UTMZ18N to DTM_hill
crs(DTM_hill_UTMZ18N_HARV)
crs(DTM_hill_HARV)

#compare attributes of DTM_hill_UTMZ18N to DTM_hill
extent(DTM_hill_UTMZ18N_HARV)
extent(DTM_hill_HARV)

```

Notice in the output above that the `CRS` of `DTM_hill_UTMZ18N_HARV` is now UTM, meters. 
However, values in the extent of `DTM_hillUTMZ18N_HARV` are different from 
`DTM_hill_HARV`.

Why do you think this is?  

<i class="fa fa-star"></i> **Data Tip:** When we reproject a raster, we 
move it from one "grid" to another. Thus, we are modifying the data! Keep this 
in mind as we work with raster data. 
{: .notice}


## Deal with Raster Resolution

Let's next have a look at the resolution of our reprojected hillshade.  

```{r view-resolution}

#compare resolution
res(DTM_hill_UTMZ18N_HARV)

```

The output resolution of `DTM_hill_UTMZ18N_HARV` is 1 x 0.998. Yet, we know that
the resolution for the data should be 1m x 1m. We can tell `R` to force our newly
reprojected raster to be 1m x 1m resolution by adding a line of code (`res=`).  

``` {r reproject-assign-resolution }
#adjust the resolution 
DTM_hill_UTMZ18N_HARV <- projectRaster(DTM_hill_HARV, 
                                  crs=crs(DTM_HARV),
                                  res=1)
#view resolution
res(DTM_hill_UTMZ18N_HARV)

```

Let's plot our newly reprojected raster.

```{r plot-projected-raster }
#plot newly reprojected hillshade
plot(DTM_hill_UTMZ18N_HARV,
    col=grey(1:100/100),
    legend=F,
    main="DTM with Hillshade\n NEON Harvard Forest Field Site")

#overlay the DTM on top of the hillshade
plot(DTM_HARV,
     col=rainbow(100),
     alpha=0.4,
     add=T,
     legend=F)
```

We have now successfully draped the Digital Terrain Model on top of our
hillshade to produce a nice looking, textured map! 

<div id="challenge" markdown="1">
## Challenge: Reproject, then Plot a Digital Terrain Model 
Create a map of the 
<a href="http://www.neoninc.org/science-design/field-sites/san-joaquin-experimental-range" target="_blank" >San Joaquin Experimental Range</a>
field site using the `SJER_DSMhill_WGS84.tif` and `SJER_dsmCrop.tif` files. 

Reproject the data as necessary to make things line up!
</div>

```{r challenge-code-reprojection, echo=FALSE }

#import DTM
DSM_SJER <- raster("NEON-DS-Airborne-Remote-Sensing/SJER/DSM/SJER_dsmcrop.tif")
#import DTM hillshade
DSM_hill_SJER_WGS <- 
  raster("NEON-DS-Airborne-Remote-Sensing/SJER/DSM/SJER_DSMhill_WGS84.tif")

#reproject raster 
DTM_hill_UTMZ18N_SJER <- projectRaster(DSM_hill_SJER_WGS, 
                                  crs=crs(DSM_SJER),
                                  res=1)
#plot hillshade using a grayscale color ramp 
plot(DTM_hill_UTMZ18N_SJER,
    col=grey(1:100/100),
    legend=F,
    main="DSM with Hillshade\n NEON SJER Field Site")

#overlay the DSM on top of the hillshade
plot(DSM_SJER,
     col=terrain.colors(10),
     alpha=0.4,
     add=T,
     legend=F)

```

<div id="challenge" markdown="1">
If you completed the San Joaquin plotting challenge in the
[Plot Raster Data in R]({{ site.baseurl}}/R/Plot-Rasters-In-R/) 
lesson, how does the map you just created compare to that map? 
</div>

```{r challenge-code-reprojection2, echo=FALSE }
#The maps look identical. Which is what they should be as the only difference
# is this one was reprojected from WGS84 to UTM prior ot plotting.  
```