---
layout: post
title: "Lesson 02: Reproject Rasters in R"
date:   2015-10-27
authors: [Jason Williams, Jeff Hollister, Kristina Riemer, Mike Smorul, Zack Brym, Leah Wasser]
contributors: [Test Human]
packagesLibraries: [raster, rgdal]
dateCreated:  2015-10-23
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
category: spatio-temporal-workshop
tags: [module-1]
mainTag: GIS-Spatial-Data
description: "This lesson explains how to reproject a raster in R using the projectRaster
function in the raster package."
code1: 
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: http://www.neoninc.org
permalink: /R/Reproject-Raster-In-R.R
comments: false
---

{% include _toc.html %}

##About


**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

###Goals / Objectives

After completing this activity, you will know:

* How to reproject a  raster in R

###Things You'll Need To Complete This Lesson

###R Libraries to Install:

* **raster:** `install.packages("raster")`
* **rgdal:** `install.packages("rgdal")`

####Tools To Install

Please be sure you have the most current version of `R` and preferably
R studio to write your code.


####Data to Download

Download the workshop data:

* <a href="http://files.figshare.com/2387965/NEON_RemoteSensing.zip" class="btn btn-success"> DOWNLOAD Sample NEON Raster Data Derived from LiDAR over Harvard
Forest and SJER Field Sites</a>

The LiDAR and imagery data used to create the rasters in this dataset were 
collected over the Harvard and San Joaquin field sites 
and processed at <a href="http://www.neoninc.org" target="_blank" >NEON </a> 
headquarters. The entire dataset can be accessed by request from the NEON website. 

####Recommended Pre-Lesson Reading

* <a href="http://cran.r-project.org/web/packages/raster/raster.pdf" target="_blank">
Read more about the `raster` package in R.</a>

</div>

#Manipulating Rasters in R

We often want to overlay two or more rasters on top of each other, and perform
calculations to create a new output raster. For example, we might have a lidar
`Digital Surface Model (DSM)` that tells us the elevation at the top of the earths surface.
This means that it is the elevation at the tops of trees, buildings and all other
objects on the earth. In comparison, a `Digital Terrain Model (DTM)` or `Digital Elevation
Model (DEM)` contains elevation data for the actual ground (with trees, buildings and 
other objects removed). 

In ecology, we are often interested in measuring the heights of trees and so the
raster that we really want is the `difference` between the `DSM` and the `DTM`.
This data product is often referred to as a `Canopy Height Model (CHM)` and represents
the actual height of trees, buildings etc above the ground.

We can calculate this difference by subtracting the two rasters in `R`. However
if our rasters get large, we can calculate the difference more efficiently using
the `overlay` function. 

First, let's create and plot a `CHM` by subtracting two rasters.

```{r load-libraries }

#load raster package
library(raster)
```

Let's create an overlay map of the Digital Terrain Model.

```{r import-DTM-hillshade }
#import DTM
DTM <- raster("NEON_RemoteSensing/HARV/DTM/HARV_dtmcrop.tif")
#import DTM hillshade
DTM_hill <- raster("NEON_RemoteSensing/HARV/DTM/HARV_DTMhill_WGS84.tif")

#NOTE: this should use color brewer for consistency across lessons
#	Plot hillshade using a grayscale color ramp 
plot(DTM_hill,
    col=grey(1:100/100),
    legend=F,
    main="NEON Hillshade - DTM\n Harvard Forest")

#overlay the DSM on top of the hillshade
plot(DTM,
     col=rainbow(100),
     alpha=0.4,
     add=T,
     legend=F)

```

So what happened? Our hillshade plotted just fine but when we tried to overlay
the DTM raster on top, nothing happened? A likely culprit for things not lining
up is the `CRS` or `Coordinate Reference System`. Let's explore our data.

```{r explore-crs }

#view crs for DTM
crs(DTM)

#view crs for hillshade
crs(DTM_hill)

```

#Reproject Rasters

When things don't line up, it is often do to differences in CRS. In this case,
our DTM is in UTM zone north. However our Hillshade is in geographic WGS84
(latitude and longitude). 

We can use the `projectRaster` function to reproject a raster as follows

`projectRaster(RasterObjects,CRSToReprojectTo)`

Keep in mind that reprojection only works when you first have a DEFINED CRS for
the raster object that you want to reproject. In this case, we do - we know that 
because CRS(DTM_hill)

```{r reproject-raster }

#reproject to WGS84
DTM_hill_UTMZ18N <- projectRaster(DTM_hill, crs=crs(DTM))

DTM_hill_UTMZ18N
#note: in this case we know that the resolution fo the data should be 1 m. We can 
#assign that in the reprojection as well
DTM_hill_UTMZ18N <- projectRaster(DTM_hill, 
                                  crs=crs(DTM),
                                  res=1)
DTM_hill_UTMZ18N

```

#A note about reprojection

Note that when you are reprojecting a raster, you are moving it from on "Grid" to
another. Thus you are modifying the data! Keep this in mind as you work with raster
data.

Once we have reprojected the raster, we can try to plot again!

```{r plot-projected-raster }

#plot newly reprojected hillshade
plot(DTM_hill_UTMZ18N,
    col=grey(1:100/100),
    legend=F,
    main="NEON Hillshade - DTM\n Harvard Forest")

#overlay the DSM on top of the hillshade
plot(DTM,
     col=rainbow(100),
     alpha=0.4,
     add=T,
     legend=F)
```

#challenge?

