---
layout: post
title: "Lesson 07: Extract NDVI Summary Values from a Raster Time Series"
date:   2015-10-22
authors: [Leah Wasser, Kristina Riemer, Zack Bryn, Jason Williams, Jeff Hollister, 
Mike Smorul]
contributors: [Test Human]
packagesLibraries: [raster, rgdal]
dateCreated:  2014-11-26
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
category: time-series-workshop
tags: [raster-ts-wrksp, raster]
mainTag: raster-ts-wrksp
description: "This lesson will explore a way to extract NDVI index values from a
raster time series in R and plot them using GGPLOT. Methods learned in this lesson could be applied to any 
raster time series"
code1: 
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: http://www.neoninc.org
permalink: /R/Extract-NDVI-From-Rasters-In-R/
comments: false
---

{% include _toc.html %}

##About
This lesson will explore the functions and libraries needed to work with time series
rasters in `R`. 

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

###Goals / Objectives

After completing this activity, you will know:

* How to extract summary pixel values from a raster.
* 

###Things You'll Need To Complete This Lesson

###R Libraries to Install:

* **raster:** `install.packages("raster")`
* **rgdal:** `install.packages("rgdal")`


####Tools To Install

Please be sure you have the most current version of `R` and preferably
R studio to write your code.


####Data to Download

Download the raster files for the Harvard Forest dataset:

<a href="http://files.figshare.com/2434040/NEON_RemoteSensing.zip" class="btn btn-success"> DOWNLOAD Sample NEON Airborne Observation Platform Raster Data</a> 

The LiDAR and imagery data used to create the rasters in this dataset were 
collected over the <a href="http://www.neoninc.org/science-design/field-sites/harvard-forest" target="_blank" >Harvard</a> and 
<a href="http://www.neoninc.org/science-design/field-sites/san-joaquin-experimental-range" target="_blank" >San Joaquin</a> field sites 
and processed at <a href="http://www.neoninc.org" target="_blank" >NEON </a> 
headquarters. The entire dataset can be accessed by request from the 
<a href="http://www.neoninc.org/data-resources/get-data/airborne-data" target="_blank"> NEON 
website.</a>  

####Recommended Pre-Lesson Reading

* <a href="http://cran.r-project.org/web/packages/raster/raster.pdf" target="_blank">
Read more about the `raster` package in R.</a>

####Raster Lesson Series 
This lesson is a part of a series of raster data in R lessons:

* [Lesson 00 - Intro to Raster Data in R]({{ site.baseurl}}/R/Introduction-to-Raster-Data-In-R/)
* [Lesson 01 - Plot Raster Data in R]({{ site.baseurl}}/R/Plot-Rasters-In-R/)
* [Lesson 02 - Reproject Raster Data in R]({{ site.baseurl}}/R/Reproject-Raster-In-R/)
* [Lesson 03 - Raster Calculations in R]({{ site.baseurl}}/R/Raster-Calculations-In-R/)
* [Lesson 04 - Work With Multi-Band Rasters - Images in R]({{ site.baseurl}}/R/Multi-Band-Rasters-In-R/)
* [Lesson 05 - Raster Time Series Data in R]({{ site.baseurl}}/R/Raster-Times-Series-Data-In-R/)
* [Lesson 06 - Plot Raster Time Series Data in R Using RasterVis and LevelPlot]({{ site.baseurl}}/R/Plot-Raster-Times-Series-Data-In-R/)
* [Lesson 07- Extract NDVI Summary Values from a Raster Time Series]({{ site.baseurl}}/R/Extract-NDVI-From-Rasters-In-R/)
</div>

```{r load-libraries }

library(raster)
library(rgdal)
library(ggplot2)

```

#About the Time Series Data

In this lesson, we are working with the same set of rasters that we used in 
[{{ site.baseurl}} /R/Raster-Times-Series-Data-In-R/ ](Lesson 05 - Time Series Rasters in R), 
that were derived from the Landsat satellite - in `GeoTiff` format. Each
raster covers the <a href="http://www.neoninc.org/science-design/field-sites/harvard-forest" target="_blank">NEON Harvard Forest field site</a>.


##Getting Started 

To begin, we will use the same raster stack that we used in the previous lesson.


```{r import-ndvi-rasters }

# Create list of NDVI file paths
NDVI_path_HARV <- "Landsat_NDVI/HARV/2011/ndvi"
all_NDVI_HARV <- list.files(NDVI_path_HARV, full.names = TRUE, pattern = ".tif$")

#view list - note that the full path (relative to our working directory)
#is included
all_NDVI_HARV

# Create a time series raster stack
NDVI_stack_HARV <- stack(all_NDVI_HARV)

```

Once we have created a raster stack, we can efficiently work with our time series.

#Calculate Average NDVI

Our output goal is a dataframe that contains a single, mean NDVI value for each
raster in our time series. We can calculate the mean for each raster using the
`cellStats` function. This produces a numeric array of values.

```{r calculate-avg-NDVI }
#calculate mean NDVI for each raster
avg_NDVI_HARV <- cellStats(NDVI_stack_HARV,mean)

#convert output to data.frame
avg_NDVI_HARV <- as.data.frame(avg_NDVI_HARV)

#We can do the above two steps with one line of code
avg_NDVI_HARV <- as.data.frame(cellStats(NDVI_stack_HARV,mean))

avg_NDVI_HARV
```

Let's explore our output `data.frame`. Each row has a name corresponding with the
raster name in our stack

```{r view-dataframe-output }
#view row names for data frame
row.names(avg_NDVI_HARV)

#view the first value in the data.frame
avg_NDVI_HARV[1,1]

#view column names
names(avg_NDVI_HARV)

#rename the NDVI column
names(avg_NDVI_HARV) <- "meanNDVI"

#view cleaned column names
names(avg_NDVI_HARV)


```

We are only working with one site now, however we might want to compare several
sites worth of data. To allow us to do this, let's add a column to our `data.frame`
called "site". We can populate this with the site name.

```{r insert-site-name}

#add a site column to our data
avg_NDVI_HARV$site <- "HARV"

```



#Extract Julian Day from row.names

We'd like to produce a plot where Julian Days is on the X axis and NDVI is on the 
Y axis. To do this, we'll need a column that is populated with Julian Day values
for each NDVI value. We can do this using several approaches. For this lesson, 
e will use `gsub` to replace the `X` and the `_HARV_ndvi_crop`.

```{r extract-julian-day }

#note the use of | is equivalent to "or". this allows us to search for more than
#one pattern in our text strings
julianDays <- gsub(pattern = "X|_HARV_ndvi_crop", #the pattern to find 
            x = row.names(avg_NDVI_HARV), #the object containing the strings
            replacement = "") #what to replace each instance of the pattern with

#make sure output looks ok
head(julianDays)

#add julianDay values as a column in the dataframe
avg_NDVI_HARV$julianDay <- julianDays

#what class is the new column
class(avg_NDVI_HARV$julianDay)

#convert julian days into a time class
#avg_NDVI_HARV$julianDay <- as.POSIXlt(avg_NDVI_HARV$julianDay, format="%j")

#the code above does weird things. for one i think jday is 0 based indexing
#but also it seems to want a year and defaults to 2015. i am not sure how to force 
#it to 2011 which is when the data were collected
#being lazy and useing int for the time being.
avg_NDVI_HARV$julianDay <- as.integer(avg_NDVI_HARV$julianDay)

# Parse that character vector
#strptime(date_info, "%Y %j")
```

#Scale Data
The last thing that we need to do it so scale our data. Valid NDVI values range 
between 0-1. The metadata included a `ScaleFactor`. NDVI data are scaled by a 
factor of 10,000. A scale factor is commonly used in larger datasets. Storing decimal
places actually consumes more space and creates larger file sizes compared to 
storing integer values.

```{r scale-data }

#scale data by 10,000
avg_NDVI_HARV$meanNDVI <- avg_NDVI_HARV$meanNDVI / 10000
#view output
head(avg_NDVI_HARV)

```

#Plot NDVI Using GGPLOT

We now have a clean `data.frame` with properly scaled NDVI and julian days.
Let's plot our data.

```{r plot-data }

#plot NDVI
ggplot(avg_NDVI_HARV, aes(julianDay, meanNDVI)) +
  geom_point(size=4,colour = "blue") + 
  ggtitle("NDVI for HARV 2011\nLandsat Derived") +
  xlab("Julian Days") + ylab("Mean NDVI") +
  theme(text = element_text(size=20))


```

> ##Challenges
>
> 1. Customize your final ggplot output. Change the color of the data points
> 2. Calculate average NDVI for all rasters in the Landsat_NDVI/SJER/2013/ndvi
> directory. Then create a plot that renders BOTH sites as a comparison.



```{r challenge-answers, echo=FALSE }


# Create list of NDVI file paths
NDVI_path_SJER <- "Landsat_NDVI/SJER/2011/ndvi"
all_NDVI_SJER <- list.files(NDVI_path_SJER, full.names = TRUE, pattern = ".tif$")

# Create a time series raster stack
NDVI_stack_SJER <- stack(all_NDVI_SJER)


#Calculate Mean, Scale Data, convert to data.frame
avg_NDVI_SJER <- as.data.frame(cellStats(NDVI_stack_SJER,mean)/10000)

#rename NDVI column
names(avg_NDVI_SJER) <- "meanNDVI"


#add a site column to our data
avg_NDVI_SJER$site <- "SJER"


#Create Julian Day Column 
julianDays_SJER <- gsub(pattern = "X|_SJER_ndvi_crop", #the pattern to find 
            x = row.names(avg_NDVI_SJER), #the object containing the strings
            replacement = "") #what to replace each instance of the pattern with

#add julianDay values as a column in the dataframe
avg_NDVI_SJER$julianDay <- as.integer(julianDays_SJER)

#Merge Data Frames
ndvi_HARV_SJER <- rbind(avg_NDVI_HARV,avg_NDVI_SJER)  
  
#plot NDVI
#make it look prettier
ggplot(ndvi_HARV_SJER, aes(julianDay, meanNDVI, colour=site)) +
  geom_point(size=4,aes(group=site)) + 
  geom_line(aes(group=site)) +
  ggtitle("Landsat Derived NDVI - 2011\nNEON Harvard Forest vs San Joachim Field Sites") +
  xlab("Julian Days") + ylab("Mean NDVI") +
  theme(text = element_text(size=20))

```

