---
layout: post
title: "Lesson 07: Extract NDVI"
date:   2015-10-22
authors: [Leah Wasser, Kristina Riemer, Zack Bryn, Jason Williams, Jeff Hollister, 
Mike Smorul]
contributors: [Test Human]
packagesLibraries: [raster, rgdal, rasterVis]
dateCreated:  2014-11-26
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
category: time-series-workshop
tags: [module-1]
mainTag: GIS-Spatial-Data
description: "This lesson covers how to improve plotting output using the rasterVis
library in R. Specifically it covers using levelplot, and adding meaningful custom 
names to bands within a RasterStack."
code1: 
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: http://www.neoninc.org
permalink: /R/Extract-NDVI-From-Rasters-In-R/
comments: false
---

{% include _toc.html %}

##About
This lesson will explore the functions and libraries needed to work with time series
rasters in `R`. 

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

###Goals / Objectives

After completing this activity, you will know:

* How to assign custom names to bands in a RasterStack for prettier plotting.
* Advanced plotting using `rasterVis` library and `levelplot`

###Things You'll Need To Complete This Lesson

###R Libraries to Install:

* **raster:** `install.packages("raster")`
* **rgdal:** `install.packages("rgdal")`
* **rasterVis:** `install.packages("rasterVis")`

Note: the `rasterVis` library can be used to create nicer plots of raster time
series data! <a href="https://cran.r-project.org/web/packages/rasterVis/rasterVis.pdf"
target="_blank">Learn More about the rasterVis library</a>

####Tools To Install

Please be sure you have the most current version of `R` and preferably
R studio to write your code.


####Data to Download

Download the raster files for the Harvard Forest dataset:

<a href="http://files.figshare.com/2434040/NEON_RemoteSensing.zip" class="btn btn-success"> DOWNLOAD Sample NEON Airborne Observation Platform Raster Data</a> 

The LiDAR and imagery data used to create the rasters in this dataset were 
collected over the <a href="http://www.neoninc.org/science-design/field-sites/harvard-forest" target="_blank" >Harvard</a> and 
<a href="http://www.neoninc.org/science-design/field-sites/san-joaquin-experimental-range" target="_blank" >San Joaquin</a> field sites 
and processed at <a href="http://www.neoninc.org" target="_blank" >NEON </a> 
headquarters. The entire dataset can be accessed by request from the 
<a href="http://www.neoninc.org/data-resources/get-data/airborne-data" target="_blank"> NEON 
website.</a>  

####Recommended Pre-Lesson Reading

* <a href="http://cran.r-project.org/web/packages/raster/raster.pdf" target="_blank">
Read more about the `raster` package in R.</a>

####Raster Lesson Series 
This lesson is a part of a series of raster data in R lessons:

* [Lesson 00 - Intro to Raster Data in R]({{ site.baseurl}}/R/Introduction-to-Raster-Data-In-R/)
* [Lesson 01 - Plot Raster Data in R]({{ site.baseurl}}/R/Plot-Rasters-In-R/)
* [Lesson 02 - Reproject Raster Data in R]({{ site.baseurl}}/R/Reproject-Raster-In-R/)
* [Lesson 03 - Raster Calculations in R]({{ site.baseurl}}/R/Raster-Calculations-In-R/)
* [Lesson 04 - Work With Multi-Band Rasters - Images in R]({{ site.baseurl}}/R/Multi-Band-Rasters-In-R/)
* [Lesson 05 - Raster Time Series Data in R]({{ site.baseurl}}/R/Raster-Times-Series-Data-In-R/)
* [Lesson 06 - Plot Raster Time Series Data in R Using RasterVis and LevelPlot]({{ site.baseurl}}/R/Plot-Raster-Times-Series-Data-In-R/)
</div>

```{r load-libraries }

library(raster)
library(rgdal)

```

#About the Time Series Data

In this lesson, we are working with the same set of rasters that we used in 
[{{ site.baseurl}} /R/Raster-Times-Series-Data-In-R/ ](Lesson 05 - Time Series Rasters in R), 
that were derived from the Landsat satellite - in `GeoTiff` format. Each
raster covers the <a href="http://www.neoninc.org/science-design/field-sites/harvard-forest" target="_blank">NEON Harvard Forest field site</a>.


##Getting Started 

To begin, we will use the same raster stack that we used in the previous lesson.


```{r import-ndvi-rasters }

# Create list of NDVI file paths
NDVI_path <- "Landsat_NDVI/HARV/2011/ndvi"
all_NDVI <- list.files(NDVI_path, full.names = TRUE, pattern = ".tif$")

#view list - note that the full path (relative to our working directory)
#is included
all_NDVI

# Create a time series raster stack
NDVI_stack <- stack(all_NDVI)

```

```{r }

#create data frame, calculate NDVI
ndvi.df.HARV <- as.data.frame(matrix(-999, ncol = 2, nrow = length(all_NDVI)))

colnames(ndvi.df.HARV) <- c("julianDays", "meanNDVI")

i <- 0

for (aRaster in all_NDVI){
  i=i+1
  #open raster
  oneRaster <- raster(aRaster)
  
  #calculate the mean of each
  ndvi.df.HARV$meanNDVI[i] <- cellStats(oneRaster,mean) 
  
  #grab julian days
  ndvi.df.HARV$julianDays[i] <- substr(aRaster,nchar(aRaster)-21,nchar(aRaster)-19)
}

ndvi.df.HARV$yr <- as.integer(2011)
ndvi.df.HARV$site <- "SJER"

#plot NDVI
ggplot(ndvi.df.HARV, aes(julianDays, meanNDVI)) +
  geom_point(size=4,colour = "blue") + 
  ggtitle("NDVI for HARV 2011\nLandsat Derived") +
  xlab("Julian Days") + ylab("Mean NDVI") +
  theme(text = element_text(size=20))


```



