---
layout: post
title: "Lesson 07: Extract NDVI Summary Values from a Raster Time Series"
date:   2015-10-22
authors: [Leah Wasser, Kristina Riemer, Zack Bryn, Jason Williams, Jeff
Hollister,  Mike Smorul]
contributors: [Megan A. Jones]
packagesLibraries: [raster, rgdal]
dateCreated:  2014-11-26
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
category: time-series-workshop
tags: [raster-ts-wrksp, raster]
mainTag: raster-ts-wrksp
description: "This lesson will explore a way to extract NDVI values from a
raster time series in `R` and plot them using `ggplot`. Methods learned in this 
lesson could be applied to any raster time series set."
code1: SR07-Extract-NDVI-From-Rasters-in-R.R
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink: http://www.neoninc.org
permalink: /R/Extract-NDVI-From-Rasters-In-R/
comments: false
---

{% include _toc.html %}

##About
This lesson will explore a way to extract NDVI values from a
raster time series in `R` and plot them using `ggplot`. Methods learned in this 
lesson could be applied to any raster time series set. 

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

###Goals / Objectives

After completing this activity, you will:

* Be able to extract summary pixel values from a raster.
* Save summary values to a .csv file.
* Plot summary pixel values using `ggplot()`. 


###Things You'll Need To Complete This Lesson

Please be sure you have the most current version of `R` and, preferably,
RStudio to write your code.

###R Libraries to Install:

* **raster:** `install.packages("raster")`
* **rgdal:** `install.packages("rgdal")`

####Data to Download

<a href=" https://ndownloader.figshare.com/files/3579867" class="btn btn-success"> Download Landsat derived NDVI raster files</a> 

The imagery data used to create the rasters in this dataset were collected over
the National Ecological Observatory Network's
<a href="http://www.neoninc.org/science-design/field-sites/harvard-forest" target="_blank" >Harvard Forest</a>
and 
<a href="http://www.neoninc.org/science-design/field-sites/san-joaquin-experimental-range" target="_blank" >San Joaquin Experimental Range</a>
field sites.  
The imagery was created by the U.S. Geological Survey (USGS) using a 
<a href="http://eros.usgs.gov/#/Find_Data/Products_and_Data_Available/MSS" target="_blank" >  multispectral scanner</a>
on a <a href="http://landsat.usgs.gov" target="_blank" > Landsat Satellite </a>.
The data files are Geographic Tagged Image-File Format (GeoTIFF). 


####Setting the Working Directory
The code in this lesson assumes that you have set your working directory to the
location of the unzipped file of data downloaded above.  If you would like a
refresher on setting the working directory, please view the [Setting A Working Directory In R]({{site.baseurl}}/R/Set-Working-Directory "R Working Directory Lesson") 
lesson prior to beginning this lesson.

####Raster Lesson Series 
This lesson is a part of a series of raster data in R lessons:

* [Lesson 00 - Intro to Raster Data in R]({{ site.baseurl}}/R/Introduction-to-Raster-Data-In-R/)
* [Lesson 01 - Plot Raster Data in R]({{ site.baseurl}}/R/Plot-Rasters-In-R/)
* [Lesson 02 - Reproject Raster Data in R]({{ site.baseurl}}/R/Reproject-Raster-In-R/)
* [Lesson 03 - Raster Calculations in R]({{ site.baseurl}}/R/Raster-Calculations-In-R/)
* [Lesson 04 - Work With Multi-Band Rasters - Images in R]({{ site.baseurl}}/R/Multi-Band-Rasters-In-R/)
* [Lesson 05 - Raster Time Series Data in R]({{ site.baseurl}}/R/Raster-Times-Series-Data-In-R/)
* [Lesson 06 - Plot Raster Time Series Data in R Using RasterVis and LevelPlot]({{ site.baseurl}}/R/Plot-Raster-Times-Series-Data-In-R/)
* [Lesson 07- Extract NDVI Summary Values from a Raster Time Series]({{ site.baseurl}}/R/Extract-NDVI-From-Rasters-In-R/)

###Additional Resources

* <a href="http://cran.r-project.org/web/packages/raster/raster.pdf" target="_blank">
Read more about the `raster` package in `R`.</a>

</div>

#Why Summarize NDVI Raster Data
While raster NDVI data is great for imaging, however, as ecologists we often
want to plot a value for greenness on a plot or compare change in NDVI values 
with air temperature or precipitation.  To do this we have to get summary values
from the raster NDVI data. 

##Getting Started 
In this lesson, we are working with the same set of rasters used in the
[Raster Time Series Data in R ]({{ site.baseurl}} /R/Raster-Times-Series-Data-In-R/) and [Plot Raster Time Series Data in R Using RasterVis and Levelplot ]({{ site.baseurl}}/R/Plot-Raster-Times-Series-Data-In-R/)
lessons. This data is derived from the Landsat satellite and stored in GeoTIFF
format. Each raster covers the <a href="http://www.neoninc.org/science-design/field-sites/harvard-forest" target="_blank">NEON Harvard Forest field site</a>.  
If you do not yet have this RasterStack, please create it now. 

```{r load-libraries-data }

library(raster)
library(rgdal)
library(ggplot2)

# Create list of NDVI file paths
NDVI_path <- "Landsat_NDVI/HARV/2011/ndvi"  #assign path to object = cleaner code
all_NDVI <- list.files(NDVI_path, full.names = TRUE, pattern = ".tif$")

# Create a time series raster stack
NDVI_stack <- stack(all_NDVI)
```

#Calculate Average NDVI
Our output goal is a data frame that contains a single, mean NDVI value for each
raster in our time series. This represents the mean or average NDVI value for
the area on a given day.  

We can calculate the mean for each raster using the `cellStats` function. This
produces a numeric array of values. However, we want a data frame so we can 
convert from the array to a data frame with `as.data.frame()`.

```{r calculate-avg-NDVI }
#calculate mean NDVI for each raster
avg_NDVI_HARV <- cellStats(NDVI_stack,mean)

#convert output array to data.frame
avg_NDVI_HARV <- as.data.frame(avg_NDVI_HARV)

#To be more efficient we could do the above two steps with one line of code
#avg_NDVI_HARV <- as.data.frame(cellStats(NDVI_stack_HARV,mean))

#view data
avg_NDVI_HARV

#view only the value in row 1, column 1 of the data frame
avg_NDVI_HARV[1,1]
```

We can see that we now have a data frame with `row.names` based on the original
file name and a mean NDVI value for each. We can also select values by their
location in the data frame.

The column name `avg_NDVI_HARV` for the values is descriptive but doesn't
plot well (and it is a bit confusing to have duplicate object & column names).
Nor does "avg" expressely denote how the value is calculated. 
MeanNDVI is better. We can change the column name. 

```{r view-dataframe-output }
#view column name slot
names(avg_NDVI_HARV)

#rename the NDVI column
names(avg_NDVI_HARV) <- "meanNDVI"

#view cleaned column names
names(avg_NDVI_HARV)
```

However, by renaming we loose the information that this is at the Harvard
Forest.  While, we are only working with one site now, we
might want to compare several sites worth of data in the future. To allow us to
do this, let's add a column to our data frame called "site". We can populate this
with the site name.

```{r insert-site-name}
#add a site column to our data
avg_NDVI_HARV$site <- "HARV"

#add a "year" column to our data
avg_NDVI_HARV$year <- "2011"

#view data
head(avg_NDVI_HARV)
```

We now have data frame with each file and two values for each one `meanNDVI` and 
`site`.  

#Extract Julian Day from row.names
We'd like to produce a plot where Julian days (the numeric day of the year,
0 - 365/366) is on the x-axis and NDVI is on
the y-axis. To do this, we'll need a column that is populated with only the
Julian day value for each NDVI value. We can do this using one of several
approaches. 

For this lesson, we will use `gsub` to replace the `X` and the `_HARV_ndvi_crop`.

```{r extract-julian-day }
#note the use of the vertical bar character ( | ) is equivalent to "or". This
# allows us to search for more than one pattern in our text strings.
julianDays <- gsub(pattern = "X|_HARV_ndvi_crop", #the pattern to find 
            x = row.names(avg_NDVI_HARV), #the object containing the strings
            replacement = "") #what to replace each instance of the pattern with

#alternate format
#julianDays <- gsub("X|_HARV_ndvi_crop", "", row.names(avg_NDVI_HARV))

#make sure output looks ok
head(julianDays)

#add julianDay values as a column in the data frame
avg_NDVI_HARV$julianDay <- julianDays

#what class is the new column
class(avg_NDVI_HARV$julianDay)
```


##Create Date Column from Julian Day
Currently the Julian day is recognized as a character class value.  We'd prefer
if it was recognized by `R` as one of the data-time classes so that `R` can
interpret and format it as a date when plotting and in other functions. 

For more information on date-time classes, see the NEON Data Skills lesson 
[Convert Date & Time Data from Character Class to Date-Time Class (POSIX) in R ]({{ site.baseurl}} /R/Time-Series-Convert-Date-Time-Class-POSIX/).  

We know our data is from 2011, therefore the first day or origin for our Julian
day count is 01 January 2011.  After setting this as the origin we just need to 
add the Julian day (as an integer) to the origin date.  Since the origin data
was originally set as a Date class object, the new `Date` column is also Date 
class. 
``` {r convert-jd}
#set the origin for the julian date (1 Jan 2011)
origin<-as.Date ("2011-01-01")

#convert "julianDay" from class character to integer
avg_NDVI_HARV$julianDay <- as.integer(avg_NDVI_HARV$julianDay)

#create a date column, -1 added because origin is the 1st.  
avg_NDVI_HARV$Date<- origin + (avg_NDVI_HARV$julianDay-1)

#did it work? 
head(avg_NDVI_HARV$Date)

#What are the classes of the two columns now? 
class(avg_NDVI_HARV$Date)
class(avg_NDVI_HARV$julianDay)

```

#Scale NDVI Data
The last thing that we need to do it so scale our data. Valid NDVI values range 
between 0-1. Our data is from a min of 277 to a max of 9140, how does that work.
The metadata for this NDVI data have a scale factor: 10,000. A scale factor is
commonly used in larger datasets to remove decimals. Storing decimal places
actually consumes more space and creates larger file sizes compared to storing
integer values.

As we have a small data set of only 13 values we do not need to have the values
scaled.  

```{r scale-data }
#de-scale data by 10,000
avg_NDVI_HARV$meanNDVI <- avg_NDVI_HARV$meanNDVI / 10000

#view output
head(avg_NDVI_HARV)

```

#Challenge: NDVI for the San Joaquin Experimental Range
As ecologist we often want to compare two different sites.  Up to this point,
we've only worked with data from the Harvard Forest.  However, the National
Ecological Obseravatory Network (NEON) and Landsat collect data and remote 
sensing imagery for the <a href="http://www.neoninc.org/science-design/field-sites/san-joaquin-experimental-range" target="_blank" >San Joaquin Experimental Range (SJER) </a> located in 
California.  

In order to faciliate comparison between the Harvard Forests and SJER, create a
complementary NDVI data set for SJER based on NDVI GeoTIFF files in the
Landsat_NDVI/SJER/2011/ndvi directory. 

```{r challenge-answers,  include=TRUE, results="hide", echo=FALSE} 
# Create list of NDVI file paths
NDVI_path_SJER <- "Landsat_NDVI/SJER/2011/ndvi"
all_NDVI_SJER <- list.files(NDVI_path_SJER, full.names = TRUE, pattern = ".tif$")

# Create a time series raster stack
NDVI_stack_SJER <- stack(all_NDVI_SJER)


#Calculate Mean, Scale Data, convert to data.frame all in 1 line!
avg_NDVI_SJER <- as.data.frame(cellStats(NDVI_stack_SJER,mean)/10000)

#rename NDVI column
names(avg_NDVI_SJER) <- "meanNDVI"

#add a site column to our data
avg_NDVI_SJER$site <- "SJER"

#add a "year" column to our data
avg_NDVI_SJER$year <- "2011"

#Create Julian Day Column 
julianDays_SJER <- gsub(pattern = "X|_SJER_ndvi_crop", #the pattern to find 
            x = row.names(avg_NDVI_SJER), #the object containing the strings
            replacement = "") #what to replace each instance of the pattern with

##Create Date column based on julian day & make julianDay an integer

#set the origin for the julian date (1 Jan 2011)
origin<-as.Date ("2011-01-01")

#add julianDay values as a column in the data frame
avg_NDVI_SJER$julianDay <- as.integer(julianDays_SJER)

#create a date column, 1 once since the origin IS day 1.  
avg_NDVI_SJER$Date<- origin + (avg_NDVI_SJER$julianDay-1)

#did it work? 
avg_NDVI_SJER

```

#Plot NDVI Using ggplot
We now have a clean data frame with properly scaled NDVI and Julian days.  Let's
plot our data.  

We will use the `ggplot()` function within the `ggplot2` package for this plot. 
If you are unfamiliar with `ggplot()` or would like to learn more about plotting
in `ggplot()` see the lesson on
[Plotting Time Series with ggplot in R ]({{ site.baseurl}}/R/Time-Series-Plot-ggplot/).

```{r ggplot-data }

#plot NDVI
ggplot(avg_NDVI_HARV, aes(julianDay, meanNDVI)) +
  geom_point(size=4,colour = "blue") + 
  ggtitle("NDVI for HARV 2011\nLandsat Derived") +
  xlab("Julian Days") + ylab("Mean NDVI") +
  theme(text = element_text(size=20))

```


#Challenge: ggplot for San Joaquin Experimental Range Data
Create a plot for the SJER data. Plot the data points in a different color. 

```{r challenge-code-ggplot-data }

#plot NDVI
ggplot(avg_NDVI_SJER, aes(julianDay, meanNDVI)) +
  geom_point(size=4,colour = "darkgreen") + 
  ggtitle("NDVI for SJER 2011\nLandsat Derived") +
  xlab("Julian Day") + ylab("Mean NDVI") +
  theme(text = element_text(size=20))

```

#Compare NDVI from Two Different Sites in One Plot
Comparison of plots is often easiest when both plots are side by side. Or, even 
better, if both sets of data are plotted in the same plot. We can do this by 
binding the two data sets together.  The date frames must have the same number
of columns and exact same column names to be bound.  

``` {r merge-df-single-plot}
#Merge Data Frames
ndvi_HARV_SJER <- rbind(avg_NDVI_HARV,avg_NDVI_SJER)  
  
#plot NDVI values for both sites
ggplot(ndvi_HARV_SJER, aes(julianDay, meanNDVI, colour=site)) +
  geom_point(size=4,aes(group=site)) + 
  geom_line(aes(group=site)) +
  ggtitle("Landsat Derived NDVI - 2011\nNEON Harvard Forest vs San Joaquin") +
  xlab("Julian Day") + ylab("Mean NDVI") +
  scale_colour_manual(values=c("blue", "darkgreen")) +   #match previous plots
  theme(text = element_text(size=20))

```

#Challenge: Plot NDVI with Date
Plot the SJER and HARV data in one plot but use date, not julian day, 
for the x-axis. 

``` {r challenge-code-plot2, echo=FALSE}
#plot NDVI values for both sites
ggplot(ndvi_HARV_SJER, aes(Date, meanNDVI, colour=site)) +
  geom_point(size=4,aes(group=site)) + 
  geom_line(aes(group=site)) +
  ggtitle("Landsat Derived NDVI - 2011\n Harvard Forest vs San Joaquin") +
  xlab("Date") + ylab("Mean NDVI") +
  scale_colour_manual(values=c("blue", "darkgreen")) +   #match previous plots
  theme(text = element_text(size=20))

```

#Remove Questionable Data
As we look at these plots we see variation across the year.  However, th
pattern is interrupted by a few deep dips to an NDVI value near 0.  Is the
vegetation truely dead and gone or are these inapproriate values that should be
removed from the data.  

Let's look at the RGB images from Harvard Forest.
```{r view-all-rgb, echo=FALSE}
#open up the cropped files
rgb.allCropped <-  list.files("Landsat_NDVI/HARV/2011/RGB/", 
                              full.names=TRUE, 
                              pattern = ".tif$")
#create a layout
par(mfrow=c(4,4))

#Super efficient code
for (aFile in rgb.allCropped){
  ndvi.rastStack <- stack(aFile)
  plotRGB(ndvi.rastStack, stretch="lin")
}

#reset layout
par(mfrow=c(1,1))
```

Notice that on the same days that have very low NDVI values the images are
filled with clouds.  Therefore the low NDVI is due to no reflectance from
vegetation being recorded.  While we don't have RGBImagery for SJER, we can
guess that a similar data problem occured for the two days (in January &
September) that the NDVI values dip to near 0.  

As we don't want to create the impression that NDVI is acutally lower on those
days let's remove them and replot.  

First, we need to identify the rows that should be retained. We can do this by
identifying all values less than some threshold for the questionable data.  
Given the plots of this data a threshold of 0.1 would work. We can then use the
subset function to just keep the data that is not suspect.  

```{r remove-bad-values}
#retain only rows with meanNDVI>0.1
avg_NDVI_HARV_clean<-subset(avg_NDVI_HARV, meanNDVI>0.1)

#work?
avg_NDVI_HARV_clean$meanNDVI<0.1
```

Now we can create another plot without the suspect data. 

``` {r plot-clean-HARV}
#plot without questionable data
ggplot(avg_NDVI_HARV_clean, aes(julianDay, meanNDVI)) +
  geom_point(size=4,colour = "blue") + 
  ggtitle("NDVI for HARV 2011\nLandsat Derived") +
  xlab("Julian Days") + ylab("Mean NDVI") +
  theme(text = element_text(size=20))
```  
  
Now we can see a much cleaner annual pattern.  Notice that the scale bar has now
changed. 

#Write NDVI data to a .csv File
This summarized NDVI data is in a tabular, not raster, format. We could now
compare it to other times of time-series data to compare vegetative,
meteorological, or other types of data to the the NDVI.  However, to do this we
need to export the data into a file format that can easily be used in other
programs, used in `R` at a later time, or shared with colleagues.  A Comma
Seperated Value (.csv) file format is an excellent choice for this as it is a
common plain-text format used by many programs and unlikely to become outdated.

We can use `write.csv()` to write a specified data frame to a .csv file that is 
saved in the current working directory (can specify otherwise). 

``` {r write-csv }
#confirm data frame is the way we want it
head(avg_NDVI_HARV_clean)
```

We have all the data we want in our data frame, however, we still have row names
based on the individual raster files that the NDVI mean came from.  All the data
in the names is repeated in other columns. If we wrote to a .csv currently this
the first column would be an unnamed column.  Instead we can prevent any future
problems by removing this column prior to writing to .csv.  

``` {r drop-rownames-write-csv}
#create new df to prevent changes to avg_NDVI_HARV
NDVI_HARV_toWrite<-avg_NDVI_HARV_clean

#drop the row.names column 
row.names(NDVI_HARV_toWrite)<-NULL

#check data frame
head(NDVI_HARV_toWrite)

#create a .csv of mean NDVI values being sure to give descriptive name
#write.csv(DateFrameName, file="NewFileName")
write.csv(NDVI_HARV_toWrite, file="meanNDVI_HARV_2011.csv")
```

#Challenge: Write to .csv
Create a NDVI .csv file for the SJER that is compatable with the one we just
created for the Harvard Forest.  Be sure to inspect for questionable values 
before writing any data to a .csv file. 

``` {r challenge-code-write-sjer,  include=TRUE, results="hide", echo=FALSE}
#retain only rows with meanNDVI>0.1
avg_NDVI_SJER_clean<-subset(avg_NDVI_SJER, meanNDVI>0.1)

#create new df to prevent changes to avg_NDVI_HARV
NDVI_SJER_toWrite<-avg_NDVI_SJER_clean

#drop the row.names column 
row.names(NDVI_SJER_toWrite)<-NULL

#check data frame
head(NDVI_SJER_toWrite)

#create a .csv of mean NDVI values being sure to give descriptive name
#write.csv(DateFrameName, file="NewFileName")
write.csv(NDVI_SJER_toWrite, file="meanNDVI_SJER_2011.csv")
```
