---
title: "work-with-FIU-data"
output: html_document
---

This document begins to explore approaches to handling met station collected time
series data. For the time being we are using data from the site below.

(http://harvardforest.fas.harvard.edu:8080/exist/apps/datasets/showData.html?id=hf001)[http://harvardforest.fas.harvard.edu:8080/exist/apps/datasets/showData.html?id=hf001]
A nice overview of time series formats in R

http://www.noamross.net/blog/2014/2/10/using-times-and-dates-in-r---presentation-code.html

#Load Libraries

```{r load-libraries-date-function}

#load ggplot for plotting 
library(ggplot2)
#the scales library supports breaks and formatting in ggplot
library(scales)

#don't load strings as factors
options(stringsAsFactors = FALSE)

```

#Fix Date Function

The dates natively come in - in `chr` format. We need to strip the letters and turn
it into a time class. 

**Can we use a placeholder to tell it to ignore the letters??**

```{r create-date-fix-function}

fixDate <- function(dateField,timezone){
  # Remove Characters from the dates - 2014-08-19T00:43:22Z
  # Replace T and Z with a space
  aDateField <- gsub("T|Z", " ", dateField)
  
  #set the field to be a date field
  aDateField <- as.POSIXct(aDateField,format = "%Y-%m-%d %H:%M:%S", 
                          tz = "GMT")
  
  #list of time zones
  #https://en.wikipedia.org/wiki/List_of_tz_database_time_zones
  #convert to local time for pretty plotting
  attributes(aDateField)$tzone <- timezone
  return(aDateField)
}

fixDateTime <- function(dat,timeOne,timeTwo,tz) {
  #fix dates
  dat[,timeOne] <- fixDate(dat[,timeOne],tz)
  #if there's a second time, fix that too
  if (exists("timeTwo")) {
      dat[,timeTwo] <- fixDate(dat[,timeTwo],tz)
  }
  return(dat)
}

```

#Plot NEON data

A function that deals with both the date issue (UCT to local) and plots the data.

```{r plot-NEON-data }

plotData <- function(dat, xVar, yVar,plTitle,yaxTitle) {
  # dat= a dataframe containing neon data
  # nVar = the variable of interest to be plotted

  #Create the plot
  #create a quick plot of Barometric Pressure
  
  myPlot <- ggplot(dat,aes_string(x=xVar, y=yVar)) +
           geom_point() +
           ggtitle(plTitle) +
           theme(plot.title = element_text(lineheight=.8, face="bold",size = 20)) +
           theme(text = element_text(size=20)) +
           xlab("Start Time") + ylab(yaxTitle)

  #format x axis with dates
  myPlot + scale_x_datetime(labels = date_format("%m/%d/%y"))
  }

```

#Open and clean NEON data

```{r import-data }
#import barometric pressure data
harBP <- read.csv("HarvardForest_4_2015_9_2015/bar-pressure/NEON.D01.HARV.DP1.00004.001.00000.000.025.030.BP_30min.csv")

#convert times to posix class
harBP <- fixDateTime(harBP,"startDateTime","endDateTime",tz="America/New_York")


#plot the data
plotData(harBP,"startDateTime","staPresMean", 
         "NEON Barometric Pressure\nHarvard Forest",
         "Mean Barometric Pressure")

#import single aspirated temperature data
harTemp30 <- read.csv("HarvardForest_4_2015_9_2015/s-asp-temp/NEON.D01.HARV.DP1.00002.001.00000.000.040.030.SAAT_30min.csv")

#convert times to posix class
harTemp30 <- fixDateTime(harTemp30,"startDateTime","endDateTime",tz="America/New_York")


#plot the data
#note there are missing data for May
plotData(harTemp30,"startDateTime","tempSingleMean", 
         "NEON Single Aspirated Temperature\nHarvard Forest",
         "Mean Temperature")

########################
#import 30 minute PARPAR
harPar30 <- read.csv("HarvardForest_4_2015_9_2015/PAR/NEON.D01.HARV.DP1.00024.001.00000.000.060.030.PARPAR_30min.csv")

#convert times to posix class
harPar30 <- fixDateTime(harPar30,"startDateTime","endDateTime",tz="America/New_York")


#plot the data
#note there are missing data for May
plotData(harPar30,"startDateTime","PARMean", 
         "NEON PAR - 30 Min\nHarvard Forest",
         "Mean PAR")

```


#Imported from the harvard tower 

Note - the NEON time series is not yet complete enough to review multiple years.
Thus data were downloaded from the Harvard tower.

http://harvardforest.fas.harvard.edu:8080/exist/apps/datasets/showData.html?id=hf001

NOTE: this is not NEON data - using it to have a fuller time series.

```{r import-harvard-met-data }


#set the working dir
setwd("~/Documents/data/1_DataPortal_Workshop")

#read in 15 min average data
harMet <- read.csv(file="harvard_MetData/hf001-10-15min-m.csv")

#clean up dates
#remove the "T"
#harMet$datetime <- fixDate(harMet$datetime,"America/New_York")

# Replace T and Z with a space
harMet$datetime <- gsub("T|Z", " ", harMet$datetime)
  
#set the field to be a date field
harMet$datetime <- as.POSIXct(harMet$datetime,format = "%Y-%m-%d %H:%M", 
                          tz = "GMT")

#list of time zones
#https://en.wikipedia.org/wiki/List_of_tz_database_time_zones
#convert to local time for pretty plotting
attributes(harMet$datetime)$tzone <- "America/New_York"

#subset out some of the data - 2010-2013 
yr.09.11 <- subset(harMet, datetime >= as.POSIXct('2009-01-01 00:00') & datetime <=
as.POSIXct('2011-01-01 00:00'))

#as.Date("2006-02-01 00:00:00")
#plot Some Air Temperature Data
  
myPlot <- ggplot(yr.09.11,aes(datetime, airt)) +
           geom_point() +
           ggtitle("15 min Avg Air Temperature\nHarvard Forest") +
           theme(plot.title = element_text(lineheight=.8, face="bold",size = 20)) +
           theme(text = element_text(size=20)) +
           xlab("Time") + ylab("Mean Air Temperature")

#format x axis with dates
myPlot + scale_x_datetime(labels = date_format("%m/%d/%y"))

# Plot PAR

myPlot <- ggplot(yr.09.11,aes(datetime, parr)) +
           geom_point() +
           ggtitle("15 Min PAR\nHarvard Forest") +
           theme(plot.title = element_text(lineheight=.8, face="bold",size = 20)) +
           theme(text = element_text(size=20)) +
           xlab("Time") + ylab("Mean PAR")

#format x axis with dates
myPlot + scale_x_datetime(labels = date_format("%m/%d/%y"))
```

#Variables available as Daily Averages

```{r}

# Plot Precipitation
myPlot <- ggplot(yr.09.11,aes(datetime, prec)) +
           geom_point() +
           ggtitle("Precipitation \nHarvard Forest") +
           theme(plot.title = element_text(lineheight=.8, face="bold",size = 20)) +
           theme(text = element_text(size=20)) +
           xlab("Time") + ylab("Mean Precip")

#format x axis with dates
myPlot + scale_x_datetime(labels = date_format("%m/%d/%y"))

```


```{r read-Daily-avg }


#read in daily data
harMetDaily <- read.csv(file="harvard_MetData/hf001-06-daily-m.csv")

  
#set the field to be a date field
harMetDaily$date <- as.Date(harMetDaily$date, format = "%Y-%m-%d")

#subset out some of the data - 2010-2013 
yr.09.11_monAvg <- subset(harMetDaily, date >= as.Date('2009-01-01') & date <=
as.Date('2011-01-01'))

#as.Date("2006-02-01 00:00:00")
#plot Some Air Temperature Data
  
myPlot <- ggplot(yr.09.11_monAvg,aes(date, airt)) +
           geom_point() +
           ggtitle("Daily Air Temperature\nHarvard Forest") +
           theme(plot.title = element_text(lineheight=.8, face="bold",size = 20)) +
           theme(text = element_text(size=20)) +
           xlab("Time") + ylab("Mean Air Temperature")

#format x axis with dates
myPlot + scale_x_date(labels = date_format("%m/%d/%y"))


#plot Some Air Temperature Data
  
myPlot <- ggplot(yr.09.11_monAvg,aes(date, prec)) +
           geom_point() +
           ggtitle("Daily Precipitation\nHarvard Forest") +
           theme(plot.title = element_text(lineheight=.8, face="bold",size = 20)) +
           theme(text = element_text(size=20)) +
           xlab("Time") + ylab("Mean Air Temperature")

#format x axis with dates
myPlot + scale_x_date(labels = date_format("%m/%d/%y"))

#plot Some Air Temperature Data
  
myPlot <- ggplot(yr.09.11_monAvg,aes(date, part)) +
           geom_point() +
           ggtitle("Daily Avg Total PAR\nHarvard Forest") +
           theme(plot.title = element_text(lineheight=.8, face="bold",size = 20)) +
           theme(text = element_text(size=20)) +
           xlab("Time") + ylab("Mean Total PAR")

#format x axis with dates
myPlot + scale_x_date(labels = date_format("%m/%d/%y"))

```