---
title: "Plot Raster Data in R"
teaching: 10
exercises: 0
questions:
- ""
objectives:
- "Know how to plot a single band raster in `R`."
- "Know how to layer a raster dataset on top of a hillshade to create an elegant basemap."
keypoints:
- ""
authors: [Leah A. Wasser, Megan A. Jones, Zack Brym, Kristina Riemer, Jason Williams, Jeff Hollister,  Mike Smorul, Joseph Stachelek]
contributors: [ ]
packagesLibraries: [raster, rgdal]
dateCreated:  2015-10-23
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
categories:  [self-paced-tutorial]
tags: [R, raster, spatial-data-gis]
tutorialSeries: [raster-data-series]
mainTag: raster-data-series
description: "This tutorial explains how to plot a raster in R using R's base plot
function. It also covers how to layer a raster on top of a hillshade to produce
an eloquent map."
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink:
comments: true
---

```{r echo=FALSE}
source("../bin/chunk-options.R")
```

> ## Things You’ll Need To Complete This Tutorial
> **R Skill Level:** Intermediate - you've got the basics of `R` down.
>
> You will need the most current version of `R` and, preferably, `RStudio` loaded
on your computer to complete this tutorial.
>
> ### Install R Packages
>
> * **raster:** `install.packages("raster")`
> * **rgdal:** `install.packages("rgdal")`
>
> * [More on Packages in R - Adapted from Software Carpentry.]({{site.baseurl}}/R/Packages-In-R/)
>
> #### Download Data
>
>
> ### Reference
>
> * <a href="http://cran.r-project.org/web/packages/raster/raster.pdf" target="_blank"> Read more about the `raster` package in `R`.</a>
{: .prereq}


This tutorial reviews how to plot a raster in `R` using the `plot()`
function. It also covers how to layer a raster on top of a hillshade to produce
an eloquent map.

## Plot Raster Data in R
In this tutorial, we will plot the Digital Surface Model (DSM) raster
for the NEON Harvard Forest Field Site. We will use the `hist()` function as a
tool to explore raster values. And render categorical plots, using the `breaks` argument to get bins that are meaningful representations of our data.

We will use the `raster` and `rgdal` packages in this tutorial. If you do not
have the `DSM_HARV` object from the
[Intro To Raster In R tutorial]({{ site.baseurl }}/R/Introduction-to-Raster-Data-In-R/),
please create it now.

```{r load-libraries }
# if they are not already loaded
library(rgdal)
library(raster)

# set working directory to ensure R can find the file we wish to import
# setwd("working-dir-path-here")

# import raster
DSM_HARV <- raster("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")

```

First, let's plot our Digital Surface Model object (`DSM_HARV`) using the
`plot()` function. We add a title using the argument `main = "title"`.

```{r hist-raster }
# Plot raster object
plot(DSM_HARV,
     main = "Digital Surface Model\nNEON Harvard Forest Field Site")

```

## Plotting Data Using Breaks
We can view our data "symbolized" or colored according to ranges of values
rather than using a continuous color ramp. This is comparable to a "classified"
map. However, to assign breaks, it is useful to first explore the distribution
of the data using a histogram. The `breaks` argument in the `hist()` function
tells `R` to use fewer or more breaks or bins.

If we name the histogram, we can also view counts for each bin and assigned
break values.

```{r create-histogram-breaks }
# Plot distribution of raster values
DSMhist <- hist(DSM_HARV,
     breaks = 3,
     main = "Histogram Digital Surface Model\n NEON Harvard Forest Field Site",
     col = "wheat3",  # changes bin color
     xlab = "Elevation (m)")  # label the x-axis

# Where are breaks and how many pixels in each category?
DSMhist$breaks
DSMhist$counts
```

Warning message!? Remember, the default for the histogram is to include only a
subset of 100,000 values. We could force it to show all the pixel values or we
can use the histogram as is and figure that the sample of 100,000 values
represents our data well.

Looking at our histogram, `R` has binned out the data as follows:

* 300-350m, 350-400m, 400-450m

We can determine that most of the pixel values fall in the 350-400m range with a
few pixels falling in the lower and higher range. We could specify different
breaks, if we wished to have a different distribution of pixels in each bin.

We can use those bins to plot our raster data. We will use the
`terrain.colors()` function to create a palette of 3 colors to use in our plot.

The `breaks` argument allows us to add breaks. To specify where the breaks
occur, we use the following syntax: `breaks=c(value1, value2, value3)`.
We can include as few or many breaks as we'd like.


```{r plot-with-breaks }
# plot using breaks.
plot(DSM_HARV,
     breaks = c(300, 350, 400, 450),
     col = terrain.colors(3),
     main = "Digital Surface Model (DSM)\n NEON Harvard Forest Field Site")

```

> ## Data Tip
> Note that when we assign break values a set of 4 values will result in 3 bins of data.
{: .callout}

### Format Plot
If we need to create multiple plots using the same color palette, we can create
an `R` object (`myCol`) for the set of colors that we want to use. We can then
quickly change the palette across all plots by simply modifying the `myCol`
object.

We can label the x- and y-axes of our plot too using `xlab` and `ylab`.

```{r add-plot-title }
# Assign color to a object for repeat use/ ease of changing
myCol <- terrain.colors(3)

# Add axis labels
plot(DSM_HARV,
     breaks = c(300, 350, 400, 450),
     col = myCol,
     main = "Digital Surface Model\nNEON Harvard Forest Field Site",
     xlab = "UTM Westing Coordinate (m)",
     ylab = "UTM Northing Coordinate (m)")
```

Or we can also turn off the axes altogether.

```{r turn-off-axes }
# or we can turn off the axis altogether
plot(DSM_HARV,
     breaks = c(300, 350, 400, 450),
     col = myCol,
     main = "Digital Surface Model\n NEON Harvard Forest Field Site",
     axes = FALSE)

```


> ## Challenge: Plot Using Custom Breaks
>
> Create a plot of the Harvard Forest Digital Surface Model (DSM) that has:
>
> 1. Six classified ranges of values (break points) that are evenly divided among the range of pixel values.
> 2. Axis labels
> 3. A plot title
>
> > ## Answers
> > ``` {r challenge-code-plotting, eval=TRUE, echo=FALSE}
> > # Find min & max
> > DSM_HARV@data
> > # Pixel range & even category width
> > (416.07-305.07)/6
> > # Break every 18.5m starting at 305.07
> > # Plot with 6 categories at even intervals across the pixel value range.
> > plot(DSM_HARV,
> >      #breaks = c(305, 323.5, 342, 360.5, 379, 397.5, 417),  #manual entry
> >      breaks = seq(305, 417, by=18.5),  #define start & end, and interval
> >      col = terrain.colors (6),
> >       main = "Digital Surface Model\nNEON Harvard Forest Field Site",
> >      xlab = "UTM Westing Coordinate",
> >      ylab = "UTM Northing Coordinate")
> > ```
> {: .solution}
{: .challenge}

## Layering Rasters
We can layer a raster on top of a hillshade raster for the same area, and use a
transparency factor to created a 3-dimensional shaded effect. A
hillshade is a raster that maps the shadows and texture that you would see from
above when viewing terrain.

```{r hillshade }
# import DSM hillshade
DSM_hill_HARV <-
  raster("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_DSMhill.tif")

# plot hillshade using a grayscale color ramp that looks like shadows.
plot(DSM_hill_HARV,
    col = grey(1:100/100),  # create a color ramp of grey colors
    legend = FALSE,
    main = "Hillshade - DSM\n NEON Harvard Forest Field Site",
    axes = FALSE)

```

> ## Data Tip
> Turn off, or hide, the legend on a plot using `legend=FALSE`.
{: .callout}

We can layer another raster on top of our hillshade using by using `add = TRUE`.
Let's overlay `DSM_HARV` on top of the `hill_HARV`.

``` {r overlay-hillshade}

# plot hillshade using a grayscale color ramp that looks like shadows.
plot(DSM_hill_HARV,
    col = grey(1:100/100),  #create a color ramp of grey colors
    legend = FALSE,
    main = "DSM with Hillshade \n NEON Harvard Forest Field Site",
    axes = FALSE)

# add the DSM on top of the hillshade
plot(DSM_HARV,
     col = rainbow(100),
     alpha = 0.4,
     add = T,
     legend = F)
```

The alpha value determines how transparent the colors will be (0 being
transparent, 1 being opaque). Note that here we used the color palette
`rainbow()` instead of `terrain.color()`.

* More information in the
<a href="https://stat.ethz.ch/R-manual/R-devel/library/grDevices/html/palettes.html" target="_blank">`R` color palettes documentation</a>.

> ## Challenge: Create DTM & DSM for SJER
> 
> Use the files in the `NEON_RemoteSensing/SJER/` directory to create a Digital
Terrain Model map and Digital Surface Model map of the San Joaquin Experimental
Range field site.
> 
> Make sure to:
> 
> * include hillshade in the maps,
> * label axes on the DSM map and exclude them from the DTM map,
> * a title for the maps,
> * experiment with various alpha values and color palettes to represent the
 data.
>
> > ## Answers
> > 
> > ```{r challenge-hillshade-layering, echo=TRUE}
> > # CREATE DSM MAPS
> > # import DSM
> > DSM_SJER <- raster("data/NEON-DS-Airborne-Remote-Sensing/SJER/DSM/SJER_dsmCrop.tif")
> > # import DSM hillshade
> > DSM_hill_SJER <- raster("data/NEON-DS-Airborne-Remote-Sensing/SJER/DSM/SJER_dsmHill.tif")
> > 
> > # plot hillshade using a grayscale color ramp that looks like shadows.
> > plot(DSM_hill_SJER,
> >     col = grey(1:100/100),  #create a color ramp of grey colors
> >     legend = FALSE,
> >     main = "DSM with Hillshade\n NEON SJER Field Site",
> >     axes = FALSE)
> > 
> > # add the DSM on top of the hillshade
> > plot(DSM_SJER,
> >      col = terrain.colors(100),
> >      alpha = 0.7,
> >      add = TRUE,
> >      legend = FALSE)
> > 
> > # CREATE SJER DTM MAP
> > # import DTM
> > DTM_SJER <- raster("data/NEON-DS-Airborne-Remote-Sensing/SJER/DTM/SJER_dtmCrop.tif")
> > # import DTM hillshade
> > DTM_hill_SJER <- raster("data/NEON-DS-Airborne-Remote-Sensing/SJER/DTM/SJER_dtmHill.tif")
> > 
> > # plot hillshade using a grayscale color ramp that looks like shadows.
> > plot(DTM_hill_SJER,
> >     col = grey(1:100/100),  #create a color ramp of grey colors
> >     legend = FALSE,
> >     main = "DTM with Hillshade\n NEON SJER Field Site",
> >     axes = FALSE)
> > 
> > # add the DSM on top of the hillshade
> > plot(DTM_SJER,
> >      col = terrain.colors(100),
> >      alpha = 0.4,
> >      add = TRUE,
> >      legend = FALSE)
> > ```
> {: .solution}
{: .challenge}