---
title: "Plot Raster Data in R"
teaching: 10
exercises: 0
questions:
- ""
objectives:
- "Know how to plot a single band raster in `R`."
- "Know how to layer a raster dataset on top of a hillshade to create an elegant basemap."
keypoints:
- ""
authors: [Leah A. Wasser, Megan A. Jones, Zack Brym, Kristina Riemer, Jason Williams, Jeff Hollister,  Mike Smorul, Joseph Stachelek]
contributors: [ ]
packagesLibraries: [raster, rgdal]
dateCreated:  2015-10-23
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
categories:  [self-paced-tutorial]
tags: [R, raster, spatial-data-gis]
tutorialSeries: [raster-data-series]
mainTag: raster-data-series
description: "This tutorial explains how to plot a raster in R using R's base plot
function. It also covers how to layer a raster on top of a hillshade to produce
an eloquent map."
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink:
comments: true
---

```{r echo=FALSE}
source("../bin/chunk-options.R")
source("../setup.R")
knitr_fig_path("02-")
```

> ## Things You’ll Need To Complete This Tutorial
> **R Skill Level:** Intermediate - you've got the basics of `R` down.
>
> You will need the most current version of `R` and, preferably, `RStudio` loaded
on your computer to complete this tutorial.
>
> ### Install R Packages
>
> * **raster:** `install.packages("raster")`
> * **rgdal:** `install.packages("rgdal")`
> * **ggplot2:** `install.packages("ggplot2")` 
> * **dplyr:** `install.packages("dplyr")`
>
> * [More on Packages in R - Adapted from Software Carpentry.]({{site.baseurl}}/R/Packages-In-R/)
>
> #### Download Data
>
>
> ### Reference
>
> * <a href="http://cran.r-project.org/web/packages/raster/raster.pdf" target="_blank"> Read more about the `raster` package in `R`.</a>
{: .prereq}


This tutorial reviews how to plot a raster in `R` using the `plot()`
function. It also covers how to layer a raster on top of a hillshade to produce
an eloquent map.

## Plot Raster Data in R
In this tutorial, we will plot the Digital Surface Model (DSM) raster
for the NEON Harvard Forest Field Site. We will use the `hist()` function as a
tool to explore raster values. And render categorical plots, using the `breaks` argument to get bins that are meaningful representations of our data.

We will use the `raster` and `rgdal` packages in this tutorial. If you do not
have the `DSM_HARV` object from the
[Intro To Raster In R tutorial]({{ site.baseurl }}/R/Introduction-to-Raster-Data-In-R/),
please create it now.

```{r load-libraries }
# if they are not already loaded
library(rgdal)
library(raster)

# set working directory to ensure R can find the file we wish to import
# setwd("working-dir-path-here")

# import raster
DSM_HARV <- raster("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")

```

First, let's plot our Digital Surface Model object (`DSM_HARV`) using the
`plot()` function. We add a title using the argument `main = "title"`.

```{r ggplot-raster }
library(ggplot2)

# convert to a df for plotting in two steps,
# First, to a SpatialPointsDataFrame
DSM_HARV_pts <- rasterToPoints(DSM_HARV, spatial = TRUE)
# Then to a 'conventional' dataframe
DSM_HARV_df  <- data.frame(DSM_HARV_pts)
rm(DSM_HARV_pts, DSM_HARV)

ggplot() +
 geom_raster(data = DSM_HARV_df , aes(x = x, y = y, fill = HARV_dsmCrop)) + 
    ggtitle("Continuous Elevation Map - NEON Harvard Forest Field Site")

```

## Plotting Data Using Breaks
We can view our data "symbolized" or colored according to ranges of values
rather than using a continuous color ramp. This is comparable to a "classified"
map. However, to assign breaks, it is useful to first explore the distribution
of the data using a bar plot. 
We use `dplyr`'s mutate to combined with `cut()` to split the data into 3 bins.


```{r histogram-breaks-ggplot}
library(dplyr)

DSM_HARV_df <- DSM_HARV_df %>%
                mutate(fct_elevation = cut(HARV_dsmCrop, breaks = 3))

ggplot() +
    geom_bar(data = DSM_HARV_df, aes(fct_elevation)) +
    xlab("Elevation (m)") +
    ggtitle("Histogram Digital Surface Model - NEON Harvard Forest Field Site")

```

If we are want to know what the bins are we can ask for the unique values 
of `fct_elavation`:
```{r unique-breaks}
unique(DSM_HARV_df$fct_elevation)
```

And we can get the count of values in each bin using `dplyr`'s 
`group_by` and `summarize`:

```{r breaks-count}
DSM_HARV_df %>%
        group_by(fct_elevation) %>%
        summarize(counts = n())
```

We can customize the break points for the bins that `cut()` yields using the 
`breaks` variable in a different way. 
Lets round the breaks so that we have bins ranges of 
300-349m, 350 - 399m, and 400-450m.
To implement this we pass a numeric vector of break points instead 
of the number of breaks we want.

```{r custom-bins}
custom_bins <- c(300, 350, 400, 450)

DSM_HARV_df <- DSM_HARV_df %>%
                mutate(fct_elevation_2 = cut(HARV_dsmCrop, breaks = custom_bins))

unique(DSM_HARV_df$fct_elevation_2)

```

> ## Data Tip
> Note that when we assign break values a set of 4 values will result in 3 bins of data.
{: .callout}

And now we can plot our bar plot again, using the new bins:

```{r }
ggplot() +
    geom_bar(data = DSM_HARV_df, aes(fct_elevation_2)) +
    xlab("Elevation (m)") +
    ggtitle("Histogram Digital Surface Model - NEON Harvard Forest Field Site \nCustom Bin Ranges")
```

We can also look at the how frequent each occurence of our bins is:

```{r break-count-custom}
DSM_HARV_df %>%
        group_by(fct_elevation_2) %>%
        summarize(counts = n())

```



We can use those bins to plot our raster data:

```{r ggplot-with-breaks}

ggplot() +
 geom_raster(data = DSM_HARV_df , aes(x = x, y = y, fill = fct_elevation_2)) + 
    ggtitle("Classified Elevation Map - NEON Harvard Forest Field Site")
```


The plot above uses the default colors inside `ggplot` for raster objects. 
We can specify our own colors to make the plot look a little nicer.
`R` has a built in set of colors for plotting terrain, which are built in
to the `terrain.colors` function.
Since we have three bins, we want to use the first three terrain colors:

```{r terrain-colors}
terrain.colors(3)
```

We can see the `terrain.colors()` function returns *hex colors* - 
 each of these character strings represents a color.
To use these in our map, we pass them across using the 
 `scale_fill_manual()` function.

```{r ggplot-breaks-customcolor}

ggplot() +
 geom_raster(data = DSM_HARV_df , aes(x = x, y = y,
                                      fill = fct_elevation_2)) + 
    scale_fill_manual(values = terrain.colors(3)) + 
    ggtitle("Classified Elevation Map - NEON Harvard Forest Field Site")
```


### More Plot Formatting

If we need to create multiple plots using the same color palette, we can create
an `R` object (`my_col`) for the set of colors that we want to use. We can then
quickly change the palette across all plots by simply modifying the `my_col`
object.

We can label the x- and y-axes of our plot too using `xlab` and `ylab`.
We can also give the legend a more meaningful title by passing a value 
to the `name` option of `scale_fill_manual.`

```{r add-ggplot-labels }
# Assign color to a object for repeat use/ ease of changing
my_col <- terrain.colors(3)

ggplot() +
 geom_raster(data = DSM_HARV_df , aes(x = x, y = y,
                                      fill = fct_elevation_2)) + 
    scale_fill_manual(values = my_col, name = "Elevation") + 
    ggtitle("Classified Elevation Map - NEON Harvard Forest Field Site") +
    xlab("UTM Westing Coordinate (m)") +
    ylab("UTM Northing Coordinate (m)")
```

Or we can also turn off the axes altogether via passing `element_blank()` to
the relevant parts of the `theme()` function.

```{r turn-off-axes }
# or we can turn off the axis altogether
ggplot() +
 geom_raster(data = DSM_HARV_df , aes(x = x, y = y,
                                      fill = fct_elevation_2)) + 
    scale_fill_manual(values = my_col, name = "Elevation") + 
    ggtitle("Classified Elevation Map - NEON Harvard Forest Field Site") +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank())
```


> ## Challenge: Plot Using Custom Breaks
>
> Create a plot of the Harvard Forest Digital Surface Model (DSM) that has:
>
> 1. Six classified ranges of values (break points) that are evenly divided among the range of pixel values.
> 2. Axis labels
> 3. A plot title
>
> > ## Answers
> > ``` {r challenge-code-plotting, eval=TRUE, echo=FALSE}
> > # Find min & max
> > 
> > DSM_HARV_df <- DSM_HARV_df  %>%
> >                mutate(fct_elevation_6 = cut(HARV_dsmCrop, breaks = 6)) 
> > 
> >  my_col <- terrain.colors(6)
> > 
> > ggplot() +
> >     geom_raster(data = DSM_HARV_df , aes(x = x, y = y,
> >                                       fill = fct_elevation_6)) + 
> >     scale_fill_manual(values = my_col, name = "Elevation") + 
> >     ggtitle("Classified Elevation Map - NEON Harvard Forest Field Site") +
> >     xlab("UTM Westing Coordinate (m)") +
> >     ylab("UTM Northing Coordinate (m)")
> > ```
> {: .solution}
{: .challenge}

## Layering Rasters
We can layer a raster on top of a hillshade raster for the same area, and use a
transparency factor to created a 3-dimensional shaded effect. A
hillshade is a raster that maps the shadows and texture that you would see from
above when viewing terrain.
We will add a custom color, making the plot grey. 

```{r hillshade }
# import DSM hillshade
DSM_hill_HARV <-
  raster("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_DSMhill.tif")

# convert to a df for plotting in two steps,
# First, to a SpatialPointsDataFrame
DSM_hill_HARV_pts <- rasterToPoints(DSM_hill_HARV, spatial = TRUE)
# Then to a 'conventional' dataframe
DSM_hill_HARV_df  <- data.frame(DSM_hill_HARV_pts)
rm(DSM_hill_HARV_pts, DSM_hill_HARV)

grey_palette <- grey.colors(100, start = 0.1, end = 0.9, 
                            gamma = 2.2, alpha = 0.4)


ggplot() +
 geom_raster(data = DSM_hill_HARV_df , aes(x = x, y = y, fill = HARV_DSMhill)) + 
    scale_fill_gradientn(colors = grey_palette, name = "Elevation") +
    ggtitle("Hillshade - DSM - NEON Harvard Forest Field Site")

```

> ## Data Tip
> Turn off, or hide, the legend on a plot using by adding `theme(legend.position="none")`.
{: .callout}

We can layer another raster on top of our hillshade using by using `add = TRUE`.
Let's overlay `DSM_HARV` on top of the `hill_HARV`.

``` {r overlay-hillshade}
hill_shade <- ggplot() +
                 geom_raster(data = DSM_hill_HARV_df , 
                             aes(x = x, y = y, fill = HARV_DSMhill)
                             ) + 
                scale_fill_gradientn(colors = grey_palette, 
                                     name = "Elevation") +
                ggtitle("Hillshade - DSM - NEON Harvard Forest Field Site")

# can't use multiple scales, # this doesnt work right now
hill_shade +
    geom_raster(data = DSM_HARV_df ,
                         aes(x = x, y = y, fill = HARV_dsmCrop)
                         ) +
    scale_color_gradientn(colors = rainbow(100))
```

The alpha value determines how transparent the colors will be (0 being
transparent, 1 being opaque). Note that here we used the color palette
`rainbow()` instead of `terrain.color()`.

* More information in the
<a href="https://stat.ethz.ch/R-manual/R-devel/library/grDevices/html/palettes.html" target="_blank">`R` color palettes documentation</a>.

> ## Challenge: Create DTM & DSM for SJER
> 
> Use the files in the `NEON_RemoteSensing/SJER/` directory to create a Digital
Terrain Model map and Digital Surface Model map of the San Joaquin Experimental
Range field site.
> 
> Make sure to:
> 
> * include hillshade in the maps,
> * label axes on the DSM map and exclude them from the DTM map,
> * a title for the maps,
> * experiment with various alpha values and color palettes to represent the
 data.
>
> > ## Answers
> > 
> > ```{r challenge-hillshade-layering, echo=TRUE}
> > # CREATE DSM MAPS
> > # import DSM
> > DSM_SJER <- raster("data/NEON-DS-Airborne-Remote-Sensing/SJER/DSM/SJER_dsmCrop.tif")
> > # import DSM hillshade
> > DSM_hill_SJER <- raster("data/NEON-DS-Airborne-Remote-Sensing/SJER/DSM/SJER_dsmHill.tif")
> > 
> > # plot hillshade using a grayscale color ramp that looks like shadows.
> > plot(DSM_hill_SJER,
> >     col = grey(1:100/100),  #create a color ramp of grey colors
> >     legend = FALSE,
> >     main = "DSM with Hillshade\n NEON SJER Field Site",
> >     axes = FALSE)
> > 
> > # add the DSM on top of the hillshade
> > plot(DSM_SJER,
> >      col = terrain.colors(100),
> >      alpha = 0.7,
> >      add = TRUE,
> >      legend = FALSE)
> > 
> > # CREATE SJER DTM MAP
> > # import DTM
> > DTM_SJER <- raster("data/NEON-DS-Airborne-Remote-Sensing/SJER/DTM/SJER_dtmCrop.tif")
> > # import DTM hillshade
> > DTM_hill_SJER <- raster("data/NEON-DS-Airborne-Remote-Sensing/SJER/DTM/SJER_dtmHill.tif")
> > 
> > # plot hillshade using a grayscale color ramp that looks like shadows.
> > plot(DTM_hill_SJER,
> >     col = grey(1:100/100),  #create a color ramp of grey colors
> >     legend = FALSE,
> >     main = "DTM with Hillshade\n NEON SJER Field Site",
> >     axes = FALSE)
> > 
> > # add the DSM on top of the hillshade
> > plot(DTM_SJER,
> >      col = terrain.colors(100),
> >      alpha = 0.4,
> >      add = TRUE,
> >      legend = FALSE)
> > ```
> {: .solution}
{: .challenge}
