---
title: "Plot Raster Time Series Data in R"
teaching: 10
exercises: 0
questions:
- ""
objectives:
- "Be able to assign custom names to bands in a RasterStack for prettier plotting."
- "Understand advanced plotting of rasters using the `rasterVis` package and `levelplot`."
keypoints:
- ""
authors: [Leah A. Wasser, Megan A. Jones, Zack Brym, Kristina Riemer, Jason Williams, Jeff Hollister,  Mike Smorul, Joseph Stachelek]
contributors: [ ]
packagesLibraries: [raster, rgdal, rasterVis]
dateCreated:  2014-11-26
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
categories:  [self-paced-tutorial]
tags: [R, raster, spatial-data-gis]
mainTag: raster-data-series
tutorialSeries: [raster-data-series, raster-time-series]
description: "This tutorial covers how to efficiently and effectively plot a
stack of rasters using rasterVis package in R. Specifically it covers using
levelplot and adding meaningful, custom names to band labels in a RasterStack."
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink:
comments: true
---

```{r echo=FALSE}
source("../bin/chunk-options.R")
```

> ## Things You’ll Need To Complete This Tutorial
> **R Skill Level:** Intermediate - you've got the basics of `R` down.
> You will need the most current version of `R` and, preferably, `RStudio` loaded
> on your computer to complete this tutorial.
>
> ### Install R Packages
>
> * **raster:** `install.packages("raster")`
> * **rgdal:** `install.packages("rgdal")`
> * **rasterVis:** `install.packages("rasterVis")`
>
> * [More on Packages in R - Adapted from Software Carpentry.]({{site.baseurl}}/R/Packages-In-R/)
>
> #### Data to Download
{: .prereq}

This tutorial covers how to improve plotting output using the `rasterVis` package
in `R`. Specifically it covers using `levelplot()` and adding meaningful custom
names to bands within a `RasterStack`.

## Get Started
In this tutorial, we are working with the same set of rasters used in the
[Raster Time Series Data in R ]({{ site.baseurl }}/R/Raster-Times-Series-Data-In-R/)
tutorial. This data is derived from the Landsat satellite and stored in
`GeoTIFF` format. Each raster covers the
<a href="http://www.neonscience.org/science-design/field-sites/harvard-forest" target="_blank">NEON Harvard Forest field site</a>.

If you have not already created the RasterStack, originally created in
[Raster Time Series Data in R ]({{ site.baseurl }}/R/Raster-Times-Series-Data-In-R/),
please create it now.

```{r load-libraries-data }

library(raster)
library(rgdal)
library(rasterVis)

# Create list of NDVI file paths
all_NDVI_HARV <- list.files("data/NEON-DS-Landsat-NDVI/HARV/2011/NDVI", full.names = TRUE, pattern = ".tif$")

# Create a time series raster stack
NDVI_HARV_stack <- stack(all_NDVI_HARV)

# apply scale factor
NDVI_HARV_stack <- NDVI_HARV_stack/10000

```

## Plot Raster Time Series Data
We can use the `plot` function to plot our raster time series data.

```{r plot-time-series }
# view a histogram of all of the rasters
# nc specifies number of columns
plot(NDVI_HARV_stack,
     zlim = c(.15, 1),
     nc = 4)

```

> ## Data Tip
> The range of values for NDVI is 0-1.
> However, the data stored in our raster ranges from 0 - 10,000. If we view the
> metadata for the original .tif files, we will see a scale factor of 10,000 is
> defined.
> Multiplying values with decimal places by a factor of 10, allows the data to be
> stored in integer format (no decimals) rather than a floating point format
> (containing decimals). This keeps the file size smaller.
{: .callout}

Our plot is nice however, it's missing some key elements including, easily
readable titles. It also contains a legend that is repeated for each image. We
can use `levelplot` from the `rasterVis` package to make our plot prettier!

* <a href="http://oscarperpinan.github.io/rastervis/" target="_blank">More on
the `rasterVis` package</a>

The syntax for the `levelplot()` function is similar to that for the `plot()`
function. We use `main = "TITLE"` to add a title to the entire plot series.

```{r levelplot-time-series }
# create a `levelplot` plot
levelplot(NDVI_HARV_stack,
          main = "Landsat NDVI\nNEON Harvard Forest")

```

## Adjust the Color Ramp
Next, let's adjust the color ramp used to render the rasters. First, we
can change the red color ramp to a green one that is more visually suited to our
NDVI (greenness) data using the `colorRampPalette()` function in combination with
`colorBrewer`.

``` {r change-color-ramp }
# use colorbrewer which loads with the rasterVis package to generate
# a color ramp of yellow to green
cols <- colorRampPalette(brewer.pal(9, "YlGn"))
# create a level plot - plot
levelplot(NDVI_HARV_stack,
        main = "Landsat NDVI -- Improved Colors \nNEON Harvard Forest Field Site",
        col.regions=cols)

```

The yellow to green color ramp visually represents NDVI well given it's a
measure of greenness. Someone looking at the plot can quickly understand that
pixels that are more green, have a higher NDVI value.

* For all of the `brewer.pal` ramp names see the
<a href="http://www.datavis.ca/sasmac/brewerpal.html" target="_blank">
brewerpal page</a>.

> ## Data Tip
> Cynthia Brewer, the creater of
> ColorBrewer, offers an online tool to help choose suitable color ramps, or to
> create your own. <a href="http://colorbrewer2.org/" target="_blank">
> ColorBrewer 2.0; Color Advise for Cartography </a>
{: .callout}

## Refine Plot & Tile Labels
Next, let's label each raster in our plot with the Julian day that the raster
represents. The current names come from the band (layer names) stored in the
`RasterStack` and first part each name is the Julian day.

To create a more meaningful label we can remove the "x" and replace it with
"day" using the `gsub()` function in `R`. The syntax is as follows:
`gsub("StringToReplace", "TextToReplaceIt", Robject)`.

First let's remove "_HARV_NDVI_crop" from each label.

```{r clean-up-names }

# view names for each raster layer
names(NDVI_HARV_stack)

# use gsub to modify label names.that we'll use for the plot
rasterNames  <- gsub("X", "Day ", names(NDVI_HARV_stack))

# view Names
rasterNames

# Remove HARV_NDVI_crop from the second part of the string
rasterNames  <- gsub("_HARV_ndvi_crop", "", rasterNames)

# view names for each raster layer
rasterNames
```

> ## Data Tip
> Instead of substituting "x" and
> "_HARV_NDVI_crop" separately, we could have used use the vertical bar character
> ( | ) to replace more than one element.
> For example "X|_HARV" tells `R` to replace all instances of both "X" and "_HARV"
> in the string. Example code to remove "x" an "_HARV...":
> `gsub("X|_HARV_NDVI_crop"," |  ","rasterNames")`
{: .callout}

Once the names for each band have been reassigned, we can render our plot with
the new labels using `names.attr=rasterNames`.

```{r create-levelplot }

# use level plot to create a nice plot with one legend and a 4x4 layout.
levelplot(NDVI_HARV_stack,
          layout=c(4, 4), # create a 4x4 layout for the data
          col.regions=cols, # add a color ramp
          main = "Landsat NDVI - Julian Days \nHarvard Forest 2011",
          names.attr=rasterNames)

```

We can adjust the columns of our plot too using `layout=c(cols, rows)`. Below
we adjust the layout to be a matrix of 5 columns and 3 rows.

```{r adjust-layout }
# use level plot to create a nice plot with one legend and a 4x4 layout.
levelplot(NDVI_HARV_stack,
          layout=c(5, 3), # create a 5x3 layout for the data
          col.regions=cols, # add a color ramp
          main = "Landsat NDVI - Julian Days \nHarvard Forest 2011",
          names.attr=rasterNames)
```

Finally, `scales` allows us to modify the x and y-axis scale. Let's simply
remove the axis ticks from the plot with `scales =list(draw=FALSE)`.

```{r remove-axis-ticks }
# use level plot to create a nice plot with one legend and a 4x4 layout.
levelplot(NDVI_HARV_stack,
          layout=c(5, 3), # create a 5x3 layout for the data
          col.regions=cols, # add a color ramp
          main = "Landsat NDVI - Julian Days \nHarvard Forest 2011",
          names.attr=rasterNames,
          scales=list(draw=FALSE )) # remove axes labels & ticks
```

> ## Challenge: Divergent Color Ramps
> 
> When we used `gsub` to modify the tile labels we replaced the beginning of each
> tile title with "Day". A more descriptive name could be "Julian Day".
> 
> 1. Create a plot and label each tile "Julian Day" with the julian day value
> following.
> 2. Change the colorramp to a divergent brown to green color ramp to
> represent the data. *Hint:* Use the
> <a href="http://www.datavis.ca/sasmac/brewerpal.html" target="_blank"> brewerpal page</a>
> to help choose a color ramp.
> 
> **Questions:**
> Does having a divergent color ramp represent the data
> better than a sequential color ramp (like "YlGn")? Can you think of other data
> sets where a divergent color ramp may be best?
> 
> > ## Answers
> > 
> > ```{r challenge-code-levelplot-divergent, echo=TRUE}
> > # change Day to Julian Day
> > rasterNames  <- gsub("Day","Julian Day ", rasterNames)
> > 
> > # use level plot to create a nice plot with one legend and a 5x3 layout.
> > levelplot(NDVI_HARV_stack,
> >       layout=c(5, 3), # create a 4x3 layout for the data
> >       col.regions=colorRampPalette(brewer.pal(9, "BrBG")), # specify color
> >       main = "Landsat NDVI - Julian Days - 2011 \nNEON Harvard Forest Field Site",
> >       names.attr=rasterNames)
> > 
> > # The sequential is better than the divergent as it is more akin to the process
> > # of greening up, which starts off at one end and just keeps increasing.
> > 
> > ```
> {: .solution}
{: .challenge}
