---
title: "Plot Raster Data in R"
teaching: 20
exercises: 10
questions:
- "How can I create categorized or customized maps of raster data?"
- "How can I customize the color scheme of a raster image?"
- "How can I layer raster data in a single image?"
objectives:
- "Build customized plots for a single band raster using the `ggplot2` package."
- "Layer a raster dataset on top of a hillshade to create an elegant basemap."
keypoints:
- "Continuous data ranges can be grouped into categories using `mutate()` and `cut()`."
- "Use built-in `terrain.colors()` or set your preferred color scheme manually."
- "Layer rasters on top of one another by using the `alpha` aesthetic."
source: Rmd
---

```{r setup, echo=FALSE}
source("../bin/chunk-options.R")
source("../setup.R")
knitr_fig_path("02-")
```

```{r load-libraries, echo = FALSE, results='hide', message = FALSE, warning = FALSE}
library(raster)
library(rgdal)
library(ggplot2)
library(dplyr)
```

```{r load-data, echo = FALSE}
# Learners will have this data loaded from earlier episode
# DSM data for Harvard Forest
DSM_HARV <- raster("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_dsmCrop.tif")

DSM_HARV_df <- as.data.frame(DSM_HARV, xy = TRUE)
```

> ## Things You’ll Need To Complete This Episode
> See the [lesson homepage]({{ site.baseurl }}) for detailed information about the software,
> data, and other prerequisites you will need to work through the examples in this episode.
{: .prereq}

## Plot Raster Data in R
This episode covers how to plot a raster in R using the `ggplot2`
package with customized coloring schemes. 
It also covers how to layer a raster on top of a hillshade to produce
an eloquent map. We will continue working with the Digital Surface Model (DSM) raster
for the NEON Harvard Forest Field Site. 

## Plotting Data Using Breaks
In the previous episode, we viewed our data using a continuous color ramp. For 
clarity and visibility of the plot, we may prefer to view the data "symbolized" or colored according to ranges of values. This is comparable to a "classified"
map. To do this, we need to tell `ggplot` how many groups to break our data into, and
where those breaks should be. To make these decisions, it is useful to first explore the distribution of the data using a bar plot. To begin with, we will use `dplyr`'s `mutate()` function combined with `cut()` to split the data into 3 bins.

```{r histogram-breaks-ggplot}

DSM_HARV_df <- DSM_HARV_df %>%
                mutate(fct_elevation = cut(HARV_dsmCrop, breaks = 3))

ggplot() +
    geom_bar(data = DSM_HARV_df, aes(fct_elevation))

```

If we want to know the cutoff values for the groups, we can ask for the unique values 
of `fct_elevation`:
```{r unique-breaks}
unique(DSM_HARV_df$fct_elevation)
```

And we can get the count of values in each group using `dplyr`'s 
`group_by()` and `summarize()` functions:

```{r breaks-count}
DSM_HARV_df %>%
        group_by(fct_elevation) %>%
        summarize(counts = n())
```

We might prefer to customize the cutoff values for these groups.
Lets round the cutoff values so that we have groups for the ranges of 
300-349m, 350 - 399m, and 400-450m.
To implement this we will give `mutate()` a numeric vector of break points instead 
of the number of breaks we want.

```{r custom-bins}
custom_bins <- c(300, 350, 400, 450)

DSM_HARV_df <- DSM_HARV_df %>%
  mutate(fct_elevation_2 = cut(HARV_dsmCrop, breaks = custom_bins))

unique(DSM_HARV_df$fct_elevation_2)
```

> ## Data Tips
> Note that when we assign break values a set of 4 values will result in 3 bins of data.
>
> The bin intervals are shown using `(` to mean exclusive and `]` to mean inclusive. For example: `(305, 342]` means "from 306 through 342".
{: .callout}

And now we can plot our bar plot again, using the new groups:

```{r histogram-custom-breaks}
ggplot() +
  geom_bar(data = DSM_HARV_df, aes(fct_elevation_2))
```

And we can get the count of values in each group in the same way we did before:

```{r break-count-custom}
DSM_HARV_df %>%
  group_by(fct_elevation_2) %>%
  summarize(counts = n())
```

We can use those groups to plot our raster data, with each group being a different color:

```{r raster-with-breaks}
ggplot() +
  geom_raster(data = DSM_HARV_df , aes(x = x, y = y, fill = fct_elevation_2)) + 
  coord_quickmap()
```

The plot above uses the default colors inside `ggplot` for raster objects. 
We can specify our own colors to make the plot look a little nicer.
R has a built in set of colors for plotting terrain, which are built in
to the `terrain.colors()` function.
Since we have three bins, we want to use the first three terrain colors:

```{r terrain-colors}
terrain.colors(3)
```

The `terrain.colors()` function returns *hex colors* - 
 each of these character strings represents a color.
To use these in our map, we pass them across using the 
 `scale_fill_manual()` function.

```{r ggplot-breaks-customcolors}

ggplot() +
 geom_raster(data = DSM_HARV_df , aes(x = x, y = y,
                                      fill = fct_elevation_2)) + 
    scale_fill_manual(values = terrain.colors(3)) + 
    coord_quickmap()
```

### More Plot Formatting

If we need to create multiple plots using the same color palette, we can create
an R object (`my_col`) for the set of colors that we want to use. We can then
quickly change the palette across all plots by modifying the `my_col`
object, rather than each individual plot.

We can label the x- and y-axes of our plot too using `xlab` and `ylab`.
We can also give the legend a more meaningful title by passing a value 
to the `name` argument of the `scale_fill_manual()` function.

```{r add-ggplot-labels}

my_col <- terrain.colors(3)

ggplot() +
 geom_raster(data = DSM_HARV_df , aes(x = x, y = y,
                                      fill = fct_elevation_2)) + 
    scale_fill_manual(values = my_col, name = "Elevation") + 
    coord_quickmap()
```

Or we can also turn off the axes altogether by passing `element_blank()` to
the relevant parts of the `theme()` function.

```{r turn-off-axes}
ggplot() +
 geom_raster(data = DSM_HARV_df , aes(x = x, y = y,
                                      fill = fct_elevation_2)) + 
    scale_fill_manual(values = my_col, name = "Elevation") +
    theme(axis.title.x = element_blank(),
          axis.title.y = element_blank()) + 
    coord_quickmap()
```

> ## Challenge: Plot Using Custom Breaks
>
> Create a plot of the Harvard Forest Digital Surface Model (DSM) that has:
>
> 1. Six classified ranges of values (break points) that are evenly divided among the range of pixel values.
> 2. Axis labels.
> 3. A plot title.
>
> > ## Answers
> > ``` {r challenge-code-plotting}
> > 
> > DSM_HARV_df <- DSM_HARV_df  %>%
> >                mutate(fct_elevation_6 = cut(HARV_dsmCrop, breaks = 6)) 
> > 
> >  my_col <- terrain.colors(6)
> > 
> > ggplot() +
> >     geom_raster(data = DSM_HARV_df , aes(x = x, y = y,
> >                                       fill = fct_elevation_6)) + 
> >     scale_fill_manual(values = my_col, name = "Elevation") + 
> >     ggtitle("Classified Elevation Map - NEON Harvard Forest Field Site") +
> >     xlab("UTM Westing Coordinate (m)") +
> >     ylab("UTM Northing Coordinate (m)") + 
> >     coord_quickmap()
> > ```
> {: .solution}
{: .challenge}

## Layering Rasters

We can layer a raster on top of a hillshade raster for the same area, and use a
transparency factor to create a 3-dimensional shaded effect. A
hillshade is a raster that maps the shadows and texture that you would see from
above when viewing terrain.
We will add a custom color, making the plot grey. 

First we need to read in our DSM hillshade data and view the structure:

```{r}
DSM_hill_HARV <-
  raster("data/NEON-DS-Airborne-Remote-Sensing/HARV/DSM/HARV_DSMhill.tif")

DSM_hill_HARV
```

Next we convert it to a dataframe, so that we can plot it using `ggplot2`:

```{r}
DSM_hill_HARV_df <- as.data.frame(DSM_hill_HARV, xy = TRUE) 

str(DSM_hill_HARV_df)
```

Now we can plot the hillshade data:

```{r raster-hillshade}
ggplot() +
  geom_raster(data = DSM_hill_HARV_df,
              aes(x = x, y = y, alpha = HARV_DSMhill)) + 
  scale_alpha(range =  c(0.15, 0.65), guide = "none") + 
  coord_quickmap()
```

> ## Data Tips
> Turn off, or hide, the legend on a plot using by adding `guide = "none"` 
> to a `scale_something()` function or by setting
> `theme(legend.position="none")`.
> 
> The alpha value determines how transparent the colors will be (0 being
> transparent, 1 being opaque).
{: .callout}

We can layer another raster on top of our hillshade using adding another call to 
the `geom_raster()` function. Let's overlay `DSM_HARV` on top of the `hill_HARV`.

``` {r overlay-hillshade}
ggplot() +
  geom_raster(data = DSM_HARV_df , 
              aes(x = x, y = y, 
                  fill = HARV_dsmCrop)) + 
  geom_raster(data = DSM_hill_HARV_df, 
              aes(x = x, y = y, 
                  alpha = HARV_DSMhill)) +  
  scale_fill_viridis_c() +  
  scale_alpha(range = c(0.15, 0.65), guide = "none") +  
  ggtitle("Elevation with hillshade") +
  coord_quickmap()
```

> ## Challenge: Create DTM & DSM for SJER
> 
> Use the files in the `NEON_RemoteSensing/SJER/` directory to create a Digital
Terrain Model map and Digital Surface Model map of the San Joaquin Experimental
Range field site.
> 
> Make sure to:
> 
> * include hillshade in the maps,
> * label axes on the DSM map and exclude them from the DTM map,
> * include a title for each map,
> * experiment with various alpha values and color palettes to represent the
 data.
>
> > ## Answers
> > 
> > ```{r challenge-hillshade-layering, echo=TRUE}
> > # CREATE DSM MAPS
> > 
> > # import DSM data
> > DSM_SJER <- raster("data/NEON-DS-Airborne-Remote-Sensing/SJER/DSM/SJER_dsmCrop.tif")
> > # convert to a df for plotting
> > DSM_SJER_df <- as.data.frame(DSM_SJER, xy = TRUE)
> >
> > # import DSM hillshade
> > DSM_hill_SJER <- raster("data/NEON-DS-Airborne-Remote-Sensing/SJER/DSM/SJER_dsmHill.tif")
> > # convert to a df for plotting
> > DSM_hill_SJER_df <- as.data.frame(DSM_hill_SJER, xy = TRUE)
> > 
> > # Build Plot
> > ggplot() +
> >     geom_raster(data = DSM_SJER_df , 
> >                 aes(x = x, y = y, 
> >                      fill = SJER_dsmCrop,
> >                      alpha=0.8)
> >                 ) + 
> >     geom_raster(data = DSM_hill_SJER_df, 
> >                 aes(x = x, y = y, 
> >                   alpha = SJER_dsmHill)
> >                 ) +
> >     scale_fill_viridis_c() +
> >     guides(fill = guide_colorbar()) +
> >     scale_alpha(range = c(0.4, 0.7), guide = "none") +
> >     # remove grey background and grid lines
> >     theme_bw() + 
> >     theme(panel.grid.major = element_blank(), 
> >           panel.grid.minor = element_blank()) +
> >     xlab("UTM Westing Coordinate (m)") +
> >     ylab("UTM Northing Coordinate (m)") +
> >     ggtitle("DSM with Hillshade") +
> >     coord_quickmap()
> > 
> > # CREATE DTM MAP
> > # import DTM
> > DTM_SJER <- raster("data/NEON-DS-Airborne-Remote-Sensing/SJER/DTM/SJER_dtmCrop.tif")
> > DTM_SJER_df <- as.data.frame(DTM_SJER, xy = TRUE)
> >
> > # DTM Hillshade
> > DTM_hill_SJER <- raster("data/NEON-DS-Airborne-Remote-Sensing/SJER/DTM/SJER_dtmHill.tif")
> > DTM_hill_SJER_df <- as.data.frame(DTM_hill_SJER, xy = TRUE)
> > 
> > ggplot() +
> >     geom_raster(data = DTM_SJER_df ,
> >                 aes(x = x, y = y,
> >                      fill = SJER_dtmCrop,
> >                      alpha = 2.0)
> >                 ) +
> >     geom_raster(data = DTM_hill_SJER_df,
> >                 aes(x = x, y = y,
> >                   alpha = SJER_dtmHill)
> >                 ) +
> >     scale_fill_viridis_c() +
> >     guides(fill = guide_colorbar()) +
> >     scale_alpha(range = c(0.4, 0.7), guide = "none") +
> >     theme_bw() +
> >     theme(panel.grid.major = element_blank(), 
> >           panel.grid.minor = element_blank()) +
> >     theme(axis.title.x = element_blank(),
> >           axis.title.y = element_blank()) +
> >     ggtitle("DTM with Hillshade") +
> >     coord_quickmap()
> > ```
> {: .solution}
{: .challenge}

{% include links.md %}
