---
title: "Plot Multiple Shapefiles in R"
teaching: 10
exercises: 0
questions:
- "How to create map compositions with custom legends in base plot."
objectives:
- "Be able to plot multiple shapefiles using base R graphics."
- "Be able to apply custom symbology to spatial objects in a plot in R."
- "Be able to customize a baseplot legend in R."
keypoints:
- ""
authors: [Joseph Stachelek, Leah A. Wasser, Megan A. Jones]
contributors: [Sarah Newman]
dateCreated:  2016-02-09
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [rgdal, raster]
categories: [self-paced-tutorial]
mainTag: vector-data-series
tags: [vector-data, R, spatial-data-gis]
tutorialSeries: [vector-data-series]
description: "This tutorial provides an overview of how to create a a plot of
multiple shapefiles using base R plot. It also explores adding a legend with
custom symbols that match your plot colors and symbols."
image:
  feature: NEONCarpentryHeader_2.png
  credit: A collaboration between the National Ecological Observatory Network (NEON) and Data Carpentry
  creditlink:
comments: true
---

```{r echo=FALSE}
source("../bin/chunk-options.R")
```

> ## Things You’ll Need To Complete This Tutorial
> **R Skill Level:** Intermediate - you've got the basics of `R` down.
> You will need the most current version of `R` and, preferably, `RStudio` loaded
> on your computer to complete this tutorial.
>
> ### Install R Packages
>
> * **raster:** `install.packages("raster")`
> * **sf:** `install.packages("sf")`
>
> [More on Packages in R - Adapted from Software Carpentry.]({{site.baseurl}}/R/Packages-In-R/)
>
> ## Download Data
> * [Site layout shapefiles](https://ndownloader.figshare.com/files/3708751)
{: .prereq}

This tutorial builds upon
[the previous tutorial]({{ site.baseurl }}/R/shapefile-attributes-in-R/)
to work with shapefile attributes in `R` and explores how to plot multiple
shapefiles using base R graphics. It then covers
how to create a custom legend with colors and symbols that match your plot.

## Load the Data
To work with vector data in `R`, we can use the `sf` library. The `raster`
package also allows us to explore metadata using similar commands for both
raster and vector files.

We will import three shapefiles. The first is our `AOI` or *area of
interest* boundary polygon that we worked with in
[Open and Plot Shapefiles in R]({{site.baseurl}}/R/open-shapefiles-in-R/).
The second is a shapefile containing the location of roads and trails within the
field site. The third is a file containing the Harvard Forest Fisher tower
location. These latter two we worked with in the
[Explore Shapefile Attributes & Plot Shapefile Objects by Attribute Value in R]({{site.baseurl}}/R/shapefile-attributes-in-R/) tutorial.

```{r load-packages-data }
# load packages
library(sf)
library(raster)

# set working directory to data folder
# setwd("pathToDirHere")

# Import a polygon shapefile
aoi_boundary_HARV <- st_read("data/NEON-DS-Site-Layout-Files/HARV/HarClip_UTMZ18.shp")

# Import a line shapefile
lines_HARV <- st_read("data/NEON-DS-Site-Layout-Files/HARV/HARV_roads.shp")

# Import a point shapefile
point_HARV <- st_read("data/NEON-DS-Site-Layout-Files/HARV/HARVtower_UTM18N.shp")
```

## Plot Data

In the
[*Explore Shapefile Attributes & Plot Shapefile Objects by Attribute Value in R* tutorial]({{ site.baseurl }}/R/shapefile-attributes-in-R/)
we created a plot where we customized the width of each line in a spatial object
according to a factor level or category. To do this, we create a vector of
colors containing a color value for EACH feature in our spatial object grouped
by factor level or category.

```{r plot-unique-lines }

# view the factor levels
levels(lines_HARV$TYPE)
# create vector of line width values
lineWidth <- c(2, 4, 3, 8)[lines_HARV$TYPE]
# view vector
lineWidth

# create a color palette of 4 colors - one for each factor level
road_palette <- c("blue", "green", "grey", "purple")
road_palette
# create a vector of colors - one for each feature in our vector object
# according to its attribute value
road_colors <- c("blue", "green", "grey", "purple")[lines_HARV$TYPE]
road_colors

# create vector of line width values
lineWidth <- c(2, 4, 3, 8)[lines_HARV$TYPE]
# view vector
lineWidth

# in this case, boardwalk (the first level) is the widest.
plot(lines_HARV$geometry,
     col = road_colors,
     main = "NEON Harvard Forest Field Site\n Roads & Trails \nLine Width Varies by Type Attribute Value",
     lwd = lineWidth)

```

> ## Data Tip
> Given we have a factor with 4 levels,
> we can create a vector of numbers, each of which specifies the thickness of each
> feature in our `sf` object by factor level (category): `c(6, 4, 1, 2)[lines_HARV$TYPE]`
{: .callout}

## Add Plot Legend
In the
[previous tutorial]({{ site.baseurl }}/R/shapefile-attributes-in-R/),
we also learned how to add a basic legend to our plot.

* `bottomright`: We specify the **location** of our legend by using a default
keyword. We could also use `top`, `topright`, etc.
* `levels(objectName$attributeName)`: Label the **legend elements** using the
categories of `levels` in an attribute (e.g., levels(lines_HARV$TYPE) means use
the levels boardwalk, footpath, etc).
* `fill=`: apply unique **colors** to the boxes in our legend. `palette()` is
the default set of colors that `R` applies to all plots.

Let's add a legend to our plot.

```{r add-legend-to-plot }
plot(lines_HARV$geometry,
     col = road_colors,
     main = "NEON Harvard Forest Field Site\n Roads & Trails\n Default Legend")

# we can use the color object that we created above to color the legend objects
road_palette

# add a legend to our map
legend("bottomright",
       legend = levels(lines_HARV$TYPE),
       fill = road_palette,
       bty = "n", # turn off the legend border
       cex = .8) # decrease the font / legend size

```

However, what if we want to create a more complex plot with many shapefiles
and unique symbols that need to be represented clearly in a legend?

## Plot Multiple Vector Layers
Now, let's create a plot that combines our tower location (`point_HARV`),
site boundary (`aoi_boundary_HARV`) and roads (`lines_HARV`) spatial objects. We
will need to build a **custom legend** as well.

To begin, create a plot with the site boundary as the first layer. Then layer
the tower location and road data on top using `add = TRUE`.

```{r plot-many-shapefiles }

# Plot multiple shapefiles
plot(aoi_boundary_HARV$geometry,
     col = "grey93",
     border = "grey",
     main = "NEON Harvard Forest Field Site")

plot(lines_HARV$geometry,
     col = road_colors,
     add = TRUE)

plot(point_HARV$geometry,
     add = TRUE,
     pch = 19,
     col = "purple")

# assign plot to an object for easy modification!
plot_HARV <- recordPlot()

```


## Customize Your Legend

Next, let's build a custom legend using the symbology (the colors and symbols)
that we used to create the plot above. To do this,  we will need to build three
things:

1. A list of all "labels" (the text used to describe each element in the legend
to use in the legend.
2. A list of colors used to color each feature in our plot.
3. A list of symbols to use in the plot. NOTE: we have a combination of points,
lines and polygons in our plot. So we will need to customize our symbols!

Let's create objects for the labels, colors and symbols so we can easily reuse
them. We will start with the labels.

```{r create-custom-labels }

# create a list of all labels
labels <- c("Tower", "AOI", levels(lines_HARV$TYPE))
labels

# render plot
plot_HARV

# add a legend to our map
legend("bottomright",
       legend = labels,
       bty = "n", # turn off the legend border
       cex = .8) # decrease the font / legend size
```

Now we have a legend with the labels identified. Let's add colors to each legend
element next. We can use the vectors of colors that we created earlier to do this.

```{r add-colors }

# we have a list of colors that we used above - we can use it in the legend
road_palette

# create a list of colors to use
plot_colors <- c("purple", "grey", road_palette)
plot_colors

# render plot
plot_HARV

# add a legend to our map
legend("bottomright",
       legend = labels,
       fill = plot_colors,
       bty = "n", # turn off the legend border
       cex = .8) # decrease the font / legend size

```

Great - now we have a legend however this legend uses boxes to symbolize each
element in the plot. It might be better if the lines were symbolized as a line
and the points, symbolized as a symbol. We can customize this using
`pch=` in our legend: **16** is a point symbol, **15** is a box.

> ## Data Tip
> To view a short list of `pch` symbols,
> type `?pch` into the `R` console.
{: .callout}


```{r custom-symbols }

# create a list of pch values
# these are the symbols that will be used for each legend value
# ?pch will provide more information on values
plotSym <- c(16, 15, 15, 15, 15, 15)
plotSym

# Plot multiple shapefiles
plot_HARV

# to create a custom legend, we need to fake it
legend("bottomright",
       legend = labels,
       pch = plotSym,
       bty = "n",
       col = plot_colors,
       cex = .8)

```

Now we've added a point symbol to represent our point element in the plot. However
it might be more useful to use line symbols in our legend
rather than squares to represent the line data. We can create line symbols,
using `lty = ()`. We have a total of 6 elements in our legend:

1. A Tower Location
2. An Area of Interest (AOI)
3. and **4** Road types (levels)

The `lty` list designates, in order, which of those elements should be
designated as a line (`1`) and which should be designated as a symbol (`NA`).
Our object will thus look like `lty = c(NA, NA, 1, 1, 1, 1)`. This tells `R` to use a
line element for`the 3-6 elements in our legend only.

Once we do this, we need to **modify** our `pch` element. Each **line** element
(3-6) should be represented by a `NA` value - this tells `R` to not use a
symbol, but to instead use a line.


```{r refine-legend}
# create line object
line_legend <- c(NA, NA, 1, 1, 1, 1)
line_legend
plotSym <- c(16, 15, NA, NA, NA, NA)
plotSym

# plot multiple shapefiles
plot_HARV

# build a custom legend
legend("bottomright",
       legend = labels,
       lty = line_legend,
       pch = plotSym,
       bty = "n",
       col = plot_colors,
       cex = .8)

```


> ## Challenge: Plot Polygon by Attribute
> 
> 1. Using the `NEON-DS-Site-Layout-Files/HARV/PlotLocations_HARV.shp` shapefile,
> create a map of study plot locations, with each point colored by the soil type
> (`soilTypeOr`). How many different soil types are there at this particular field
> site? Overlay this layer on top of the `lines_HARV` layer (the roads). Create a
> custom legend that applies line symbols to lines and point symbols to the points.
> 
> 2. Modify the plot above. Tell `R` to plot each point, using a different
symbol of `pch` value. HINT: to do this, create a vector object of symbols by
factor level using the syntax described above for line width:
`c(15, 17)[lines_HARV$soilTypeOr]`. Overlay this on top of the AOI Boundary.
Create a custom legend.
> 
> 
> > ## Answers
> > 
> > ``` {r challenge-code-plot-color, results="hide", warning= FALSE, echo=TRUE}
> > ## 1
> > # open plot locations
> > plotLocations <- st_read("data/NEON-DS-Site-Layout-Files/HARV/PlotLocations_HARV.shp")
> > 
> > # how many unique soils?  Two
> > unique(plotLocations$soilTypeOr)
> > 
> > # create new color palette -- topo.colors palette
> > blueGreen <- c("blue", "darkgreen")
> > blueGreen
> > 
> > # plot roads
> > plot(lines_HARV$geometry,
> >      col=road_colors,
> >      main = "NEON Harvard Forest Field Site\n Study Plots by Soil Type\n One Symbol for All Types")
> > 
> > # plot the locations
> > plot(plotLocations$geometry,
> >      col=(blueGreen)[plotLocations$soilTypeOr],
> >      pch=18,
> >      add = TRUE)
> > 
> > # create line object
> > line_legendElement = c(NA, NA, 1, 1, 1, 1)
> > line_legendElement
> > plotSymElements <- c(18, 18, NA, NA, NA, NA)
> > plotSymElements
> > 
> > # create vector of colors
> > colorElements <- c(blueGreen, road_palette)
> > colorElements
> > 
> > # create legend
> > legend("bottomright",
> >        legend = c(levels(plotLocations$soilTypeOr), levels(lines_HARV$TYPE)),
> >        pch = plotSymElements,
> >        lty = line_legendElement,
> >        col = colorElements,
> >        bty = "n",
> >        cex = 1)
> > 
> > ## 2
> > # create vector of DIFFERENT plot symbols
> > plSymbols <- c(15, 17)[plotLocations$soilTypeOr]
> > plSymbols
> > 
> > # plot roads
> > plot(lines_HARV$geometry,
> >      col=road_colors,
> >      main = "NEON Harvard Forest Field Site\n Study Plots by Soil Type\n Different Symbols for Types")
> > 
> > # plot the locations
> > plot(plotLocations$geometry,
> >      col=(blueGreen)[plotLocations$soilTypeOr],
> >      pch=plSymbols,
> >      add = TRUE)
> > 
> > # create line object
> > line_legendElement  <- c(NA, NA, 1, 1, 1, 1)
> > line_legendElement
> > plotSymElementsMod <- c(15, 17, NA, NA, NA, NA)
> > plotSymElementsMod
> > 
> > # create vector of colors
> > colorElements <- c(blueGreen, road_palette)
> > colorElements
> > # create legend
> > legend("bottomright",
> >        legend = c(levels(plotLocations$soilTypeOr), levels(lines_HARV$TYPE)),
> >        pch = plotSymElementsMod,
> >        lty = line_legendElement,
> >        col = colorElements,
> >        bty = "n",
> >        cex = 1)
> > ```
> {: .solution}
{: .challenge}
