---
title: "Create Publication-quality Graphics"
teaching: 20
exercises: 10
questions:
- "How can I create a publication-quality graphic and customize plot parameters?"
objectives:
- "Assign custom names to bands in a RasterStack."
- "Customize raster plots using the `ggplot2` package."
keypoints:
- "Use the `theme_void()` function for a clean background to your plot."
- "Use the `element_text()` function to adjust text size, font, and position."
- "Use the `brewer.pal()` function to create a custom color palette."
- "Use the `gsub()` function to do pattern matching and replacement in text."
source: Rmd
---

```{r setup, echo=FALSE}
source("../bin/chunk-options.R")
source("../setup.R")
knitr_fig_path("13-")
```

```{r load-libraries, echo = FALSE, results='hide', message = FALSE}
library(raster)
library(rgdal)
library(ggplot2)
library(dplyr)
library(reshape)
library(RColorBrewer)
library(scales)
```

```{r load-data, echo = FALSE, results = 'hide'}
# learners will have this data loaded from the previous episode

all_NDVI_HARV <- list.files("data/NEON-DS-Landsat-NDVI/HARV/2011/NDVI", full.names = TRUE, pattern = ".tif$")

# Create a time series raster stack
NDVI_HARV_stack <- stack(all_NDVI_HARV)

# apply scale factor
NDVI_HARV_stack <- NDVI_HARV_stack/10000

# convert to a df for plotting 
NDVI_HARV_stack_df <- as.data.frame(NDVI_HARV_stack, xy = TRUE) %>%
# Then reshape data to stack all the X*_HARV_ndvi_crop columns into
# one single column called 'variable'
    melt(id.vars = c('x','y'))
```


> ## Things You’ll Need To Complete This Episode
> See the [lesson homepage]({{ site.baseurl }}) for detailed information about the software,
> data, and other prerequisites you will need to work through the examples in this episode.
{: .prereq}

This episode covers how to customize your raster plots using the `ggplot2` package
in R to create publication-quality plots. 

## Before and After

In [the previous episode]({{ site.baseurl }}/12-time-series-raster/), we learned how to plot multi-band
raster data in R using the `facet_wrap()` function. This created a separate panel in our plot
for each raster band. The plot we created together is shown below: 

```{r levelplot-time-series-before, echo = FALSE}
# code not shown, demonstration only
ggplot() +
  geom_raster(data = NDVI_HARV_stack_df , aes(x = x, y = y, fill = value)) +
  facet_wrap(~variable) +
  ggtitle("Landsat NDVI", subtitle = "NEON Harvard Forest")
```

Although this plot is informative, it isn't something we would expect to see in a journal publication. The x
and y-axis labels aren't informative. There is a lot of unnecessary gray background and the titles of
each panel don't clearly state that the number refers to the Julian day the data was collected. In this
episode, we will customize this plot above to produce a publication quality graphic. We will go through these steps iteratively. When we're done, we will have created the plot shown below.

```{r levelplot-time-series-after, echo = FALSE}
# code not shown, demonstration only

raster_names <- names(NDVI_HARV_stack)
raster_names <- gsub("_HARV_ndvi_crop", "", raster_names)
raster_names  <- gsub("X", "Day ", raster_names)
labels_names <- setNames(raster_names, unique(NDVI_HARV_stack_df$variable))
green_colors <- brewer.pal(9, "YlGn") %>%
  colorRampPalette()

ggplot() +
  geom_raster(data = NDVI_HARV_stack_df , aes(x = x, y = y, fill = value)) +
  facet_wrap(~variable, nrow = 3, ncol = 5, labeller = labeller(variable = labels_names)) +
  ggtitle("Landsat NDVI - Julian Days", subtitle = "Harvard Forest 2011") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), plot.subtitle = element_text(hjust = 0.5)) + 
  scale_fill_gradientn(name = "NDVI", colours = green_colors(20))

# cleanup
rm(raster_names, labels_names, green_colors)
```

## Adjust the Plot Theme

The first thing we will do to our plot remove the x and y-axis labels and axis ticks, as these are 
unnecessary and make our plot look messy. We can do this by setting the plot theme to `void`. 

```{r adjust-theme}
ggplot() +
  geom_raster(data = NDVI_HARV_stack_df , aes(x = x, y = y, fill = value)) +
  facet_wrap(~variable) +
  ggtitle("Landsat NDVI", subtitle = "NEON Harvard Forest") + 
  theme_void()
```

Next we will center our plot title and subtitle. We need to do this **after** the `theme_void()` layer,
because R interprets the `ggplot` layers in order. If we first tell R to center our plot title,
and then set the theme to `void`, any adjustments we've made to the plot theme will be over-written
by the `theme_void()` function. So first we make the theme `void` and then we center the title.
We center both the title and subtitle by using the `theme()` function and setting the `hjust` 
parameter to 0.5. The `hjust` parameter stands for "horizontal justification" and takes any value between 
0 and 1. A setting of 0 indicates left justification and a setting of 1 indicates right justification. 

```{r adjust-theme-2}
ggplot() +
  geom_raster(data = NDVI_HARV_stack_df , aes(x = x, y = y, fill = value)) +
  facet_wrap(~variable) +
  ggtitle("Landsat NDVI", subtitle = "NEON Harvard Forest") + 
  theme_void() + 
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

> ## Challenge
>
> Change the plot title (but not the subtitle) to bold font. You can 
> (and should!) use the help menu in RStudio or any internet resources
> to figure out how to change this setting.
> 
> > ## Answers
> > 
> > Learners can find this information in the help files for the `theme()` 
> > function. The parameter to set is called `face`. 
> > 
> > ```{r use-bold-face}
> > ggplot() +
> >   geom_raster(data = NDVI_HARV_stack_df,
> >               aes(x = x, y = y, fill = value)) +
> >   facet_wrap(~ variable) +
> >   ggtitle("Landsat NDVI", subtitle = "NEON Harvard Forest") + 
> >   theme_void() + 
> >   theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
> >         plot.subtitle = element_text(hjust = 0.5))
> > ```
> {: .solution}
{: .challenge}

## Adjust the Color Ramp

Next, let's adjust the color ramp used to render the rasters. First, we can
change the blue color ramp to a green one that is more visually suited to our
NDVI (greenness) data using the `colorRampPalette()` function in combination
with `colorBrewer`. Then we use `scale_fill_gradientn` to pass the list of
colours (here 20 different colours) to ggplot.

First we need to create a set of colors to use. We will select a set of
nine colors from the "YlGn" (yellow-green) color palette. This returns a 
set of hex color codes:

```{r}
brewer.pal(9, "YlGn")
```

Then we will pass those color codes to the `colorRampPalette` function, which
will interpolate from those colors a more nuanced color range.

```{r}
green_colors <- brewer.pal(9, "YlGn") %>%
  colorRampPalette()
```

We can 
tell the `colorRampPalette()` function how many discrete colors within this color range to
create. In our case, we will use 20 colors when we plot our graphic.

```{r change-color-ramp}
ggplot() +
  geom_raster(data = NDVI_HARV_stack_df , aes(x = x, y = y, fill = value)) +
  facet_wrap(~variable) +
  ggtitle("Landsat NDVI", subtitle = "NEON Harvard Forest") + 
  theme_void() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
    plot.subtitle = element_text(hjust = 0.5)) + 
  scale_fill_gradientn(name = "NDVI", colours = green_colors(20))
```

The yellow to green color ramp visually represents NDVI well given it's a
measure of greenness. Someone looking at the plot can quickly understand that
pixels that are more green have a higher NDVI value.

> ## Data Tip
> For all of the `brewer.pal` ramp names see the [brewerpal page](http://www.datavis.ca/sasmac/brewerpal.html).
{: .callout}

> ## Data Tip
>
> Cynthia Brewer, the creator of
> ColorBrewer, offers an online tool to help choose suitable color ramps, or to
> create your own. [ColorBrewer 2.0; Color Advise for Cartography](http://colorbrewer2.org/)
{: .callout}

## Refine Plot & Tile Labels
Next, let's label each panel in our plot with the Julian day that the 
raster data for that panel was collected. The current names come from the band
"layer names"" stored in the
`RasterStack` and the first part of each name is the Julian day.

To create a more meaningful label we can remove the "x" and replace it with
"day" using the `gsub()` function in R. The syntax is as follows:
`gsub("StringToReplace", "TextToReplaceIt", object)`.

First let's remove "_HARV_NDVI_crop" from each label to make the labels 
shorter and remove repetition. To illustrate how this works, we will first
look at the names for our `NDVI_HARV_stack` object: 

```{r}
names(NDVI_HARV_stack)
```

Now we will use the `gsub()` function to find the character string
"_HARV_ndvi_crop" and replace it with a blank string (""). We will 
assign this output to a new object (`raster_names`) and look
at that object to make sure our code is doing what we want it to.

```{r}
raster_names <- names(NDVI_HARV_stack)

raster_names <- gsub("_HARV_ndvi_crop", "", raster_names)
raster_names
```

So far so good. Now we will use `gsub()` again to replace the "X" with the word
"Day" followed by a space.

```{r}
raster_names  <- gsub("X", "Day ", raster_names)
raster_names
```

Our labels look good now. Let's reassign them to our `all_NDVI_HARV` object:
```{r}
labels_names <- setNames(raster_names, unique(NDVI_HARV_stack_df$variable))
```

Once the names for each band have been reassigned, we can render our plot with
the new labels using a`labeller`.

```{r create-levelplot }
ggplot() +
  geom_raster(data = NDVI_HARV_stack_df , aes(x = x, y = y, fill = value)) +
  facet_wrap(~variable, labeller = labeller(variable = labels_names)) +
  ggtitle("Landsat NDVI", subtitle = "NEON Harvard Forest") + 
  theme_void() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
    plot.subtitle = element_text(hjust = 0.5)) + 
  scale_fill_gradientn(name = "NDVI", colours = green_colors(20))
```


## Change Layout of Panels

We can adjust the columns of our plot by setting the number of columns `ncol`
and the number of rows `nrow` in `facet_wrap`. Let's make our plot so that
it has a width of five panels. 

```{r adjust-layout }
ggplot() +
  geom_raster(data = NDVI_HARV_stack_df , aes(x = x, y = y, fill = value)) +
  facet_wrap(~variable, ncol = 5, labeller = labeller(variable = labels_names)) +
  ggtitle("Landsat NDVI", subtitle = "NEON Harvard Forest") + 
  theme_void() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
    plot.subtitle = element_text(hjust = 0.5)) + 
  scale_fill_gradientn(name = "NDVI", colours = green_colors(20))
```

Now we have a beautiful, publication quality plot!

> ## Challenge: Divergent Color Ramps
>
> When we used the `gsub()` function to modify the tile labels we replaced the beginning of each
> tile title with "Day". A more descriptive name could be "Julian Day". Update the 
> plot above with the following changes:
>
> 1. Label each tile "Julian Day" with the julian day value
> following.
> 2. Change the color ramp to a divergent brown to green color ramp. 
>
> **Questions:**
> Does having a divergent color ramp represent the data
> better than a sequential color ramp (like "YlGn")? Can you think of other data
> sets where a divergent color ramp may be best?
>
> > ## Answers
> >
> > ```{r final-figure}
> > raster_names  <- gsub("Day","Julian Day ", raster_names)
> > labels_names <- setNames(raster_names, unique(NDVI_HARV_stack_df$variable))
> > 
> > brown_green_colors <- colorRampPalette(brewer.pal(9, "BrBG"))
> > 
> > ggplot() +
> >   geom_raster(data = NDVI_HARV_stack_df , aes(x = x, y = y, fill = value)) +
> >   facet_wrap(~variable, ncol = 5, labeller = labeller(variable = labels_names)) +
> >   ggtitle("Landsat NDVI - Julian Days", subtitle = "Harvard Forest 2011") +
> >   theme_void() +
> >   theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
> >   plot.subtitle = element_text(hjust = 0.5)) +
> >   scale_fill_gradientn(name = "NDVI", colours = brown_green_colors(20))
> > ```
> > 
> > For NDVI data, the sequential color ramp is better than the divergent as it is 
> > more akin to the process
> > of greening up, which starts off at one end and just keeps increasing.
> {: .solution}
{: .challenge}

{% include links.md %}

